---
title: "Covariate Analysis"
output: html_document
---

Load relevant data. 
```{r,  include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(ggplot2)
library(dplyr)
library(MASS)
library(scales)
library(broom)
library(gridExtra)
library(tibble)
library(stringr)
library(forestplot)
library(grid)

clin_var <- read.csv('/Users/kristinemedcalf/Desktop/Programming/Dissertation_GCA_Prediction/data/clinical_variables_cc_c1_c2.csv')
hist_var <- read.csv("/Users/kristinemedcalf/Desktop/Programming/Dissertation_GCA_Prediction/data/histological_variables_cc_c1_c2.csv")
names(hist_var)[names(hist_var) == "Lymphocytic_infiltrate_in._adventitia"] <- "Lymphocytic_infiltrate_in_adventitia"

```

Remove features with high correlation, low variance, or low usefulness: 
• Any_granulomatous_infiltrate
• PALI
• Any_lymphocytic_infiltrate
• GCA_present
• Granulomatous_infiltrate_in_media
• Hyperplasia
• Fibrosis

Granulomatous_infiltrate_in_adventitia
Granulomatous_infiltrate_in_intima
Lymphocytic_infiltrate_in_media
Lymphocytic_infiltrate_in_adventitia
Adventitia_pattern
Media_pattern
```{r}
cols_to_remove <- c("Any_granulomatous_infiltrate", "PALI", "Any_lymphocytic_infiltrate", "GCA_present", "Granulomatous_infiltrate_in_media",  "Fibrosis", "Granulomatous_infiltrate_in_adventitia", "Granulomatous_infiltrate_in_intima", "Lymphocytic_infiltrate_in_media", "Lymphocytic_infiltrate_in_adventitia",  "Lymphocytic_infiltrate_in_intima", "Barcelona Score")
hist_var <- hist_var[, setdiff(names(hist_var), cols_to_remove)]
```

Group different feature types (binary or categorical)
```{r}

# Recode inflammation severity
hist_var$Adventitia_pattern_bin <- ifelse(hist_var$Adventitia_pattern %in% c(0, 1), 0, 1)
hist_var$Media_pattern_bin      <- ifelse(hist_var$Media_pattern %in% c(0, 1), 0, 1)
hist_var$Intima_pattern_bin     <- ifelse(hist_var$Intima_pattern %in% c(0, 1), 0, 1)

hist_var$Occlusion_Grade_bin <- ifelse(
  is.na(hist_var$Occlusion_grade), NA,
  ifelse(hist_var$Occlusion_grade %in% c(0, 1), 0, 1)
)

#hist_var$Barcelona_Score_bin <- ifelse(
#  is.na(hist_var$Barcelona_score), NA,
#  ifelse(hist_var$Barcelona_score %in% c(0, 1), 0, 1)
#)

all_features <- c()
for (col in names(hist_var)) {
  if (length(levels(factor(hist_var[[col]]))) == 2) {
    all_features <- c(all_features, col)
  }
}
```

```{r}
combined <- merge(clin_var, hist_var, by = "ID")
hist_var <- as.data.frame(hist_var[, -1])

```

Steroid Exposure Models
```{r}
# Fit logistic model for each binary feature with steroid_days as predictor
steroid_model <- lapply(all_features, function(feature) {
  glm(as.formula(paste(feature, "~ steroids_days")), 
      family = binomial(link = "logit"), 
      data = combined)
})

# Name each model
names(steroid_model) <- all_features

# Summarize results
steroid_summary <- lapply(seq_along(steroid_model), function(i) {
  model <- steroid_model[[i]]
  feature <- all_features[i]
  coef_summary <- summary(model)$coefficients
  steroid_row <- rownames(coef_summary) == "steroids_days"

  data.frame(
    feature = feature,
    estimate = coef_summary[steroid_row, "Estimate"],
    lowerCI = coef_summary[steroid_row, "Estimate"] - 1.96 * coef_summary[steroid_row, "Std. Error"],
    upperCI = coef_summary[steroid_row, "Estimate"] + 1.96 * coef_summary[steroid_row, "Std. Error"],
    pvalue = coef_summary[steroid_row, "Pr(>|z|)"],
    group0 = sum(combined[[feature]] == 0, na.rm = TRUE),
    group1 = sum(combined[[feature]] == 1, na.rm = TRUE)
  )
})

# Combine and adjust p-values
steroid_summary <- do.call(rbind, steroid_summary)
steroid_summary$AdjP <- p.adjust(steroid_summary$pvalue, method = "fdr")
steroid_summary
```

Sex Models
```{r}
combined$sex_binary <- ifelse(combined$sex == 2, 1, 
                              ifelse(combined$sex == 1, 0, NA))

# Fit logistic models: feature ~ sex_binary
sex_model <- lapply(all_features, function(feature) {
  glm(as.formula(paste(feature, "~ sex_binary")), data = combined, family = binomial)
})

# Name the models by feature
names(sex_model) <- all_features

# Extract effect estimates, CIs, and p-values
sex_summary <- lapply(seq_along(sex_model), function(i) {
  model <- sex_model[[i]]
  feature <- all_features[i]

  # Get coefficient summary safely
  coef_summary <- tryCatch(summary(model)$coefficients, error = function(e) NULL)

  # Skip if model failed or sex_binary is not in the result
  if (is.null(coef_summary) || !"sex_binary" %in% rownames(coef_summary)) {
    return(NULL)
  }

  data.frame(
    feature = feature,
    OR = exp(coef_summary["sex_binary", "Estimate"]),
    lowerCI = exp(coef_summary["sex_binary", "Estimate"] - 1.96 * coef_summary["sex_binary", "Std. Error"]),
    upperCI = exp(coef_summary["sex_binary", "Estimate"] + 1.96 * coef_summary["sex_binary", "Std. Error"]),
    pvalue = coef_summary["sex_binary", "Pr(>|z|)"],
    group0 = sum(combined[[feature]] == 0, na.rm = TRUE),
    group1 = sum(combined[[feature]] == 1, na.rm = TRUE)
  )
})

# Remove any NULLs (failed models)
sex_summary <- do.call(rbind, sex_summary)
sex_summary$AdjP <- p.adjust(sex_summary$pvalue, method = "fdr")
sex_summary
```

Age Models
```{r}
# Loop over binary outcome features
age_model <- lapply(all_features, function(feature) {
  glm(as.formula(paste(feature, "~ age")), data = combined, family = binomial)
})

names(age_model) <- all_features

# Extract ORs, CIs, and p-values
age_summary <- lapply(seq_along(age_model), function(i) {
  model <- age_model[[i]]
  feature <- all_features[i]

  coef_summary <- tryCatch(summary(model)$coefficients, error = function(e) NULL)

  if (is.null(coef_summary) || !"age" %in% rownames(coef_summary)) {
    return(NULL)
  }

  est <- coef_summary["age", "Estimate"]
  se <- coef_summary["age", "Std. Error"]
  pval <- coef_summary["age", "Pr(>|z|)"]

  data.frame(
    Feature = feature,
    Predictor = "Age",
    OR = exp(est),
    LowerCI = exp(est - 1.96 * se),
    UpperCI = exp(est + 1.96 * se),
    PValue = pval,
    group0 = sum(combined[[feature]] == 0, na.rm = TRUE),
    group1 = sum(combined[[feature]] == 1, na.rm = TRUE)
  )
})

# Combine and adjust p-values
age_summary <- do.call(rbind, age_summary)
age_summary$AdjP <- p.adjust(age_summary$PValue, method = "fdr")

```


Forest Plots

Steroid Exposure Plots
```{r}
# Prepare base data frame with odds ratios
base_data <- tibble(
  mean  = exp(steroid_summary$estimate),       # OR
  lower = exp(steroid_summary$lowerCI),        # Lower 95% CI
  upper = exp(steroid_summary$upperCI),        # Upper 95% CI
  feature = steroid_summary$feature,
  pvalue_raw = steroid_summary$pvalue,
  pvalue_adj = steroid_summary$AdjP,
  estimate_ci = paste0(
    sprintf("%.2f", exp(steroid_summary$estimate)), " [",
    sprintf("%.2f", exp(steroid_summary$lowerCI)), ", ",
    sprintf("%.2f", exp(steroid_summary$upperCI)), "]"
  ),
  group0 = steroid_summary$group0,
  group1 = steroid_summary$group1

) %>%
  mutate(
    feature_clean = str_replace_all(feature, "_", " "),
    feature_clean = str_remove(feature_clean, " ?bin$"),
    feature_clean = str_to_title(feature_clean),
    
    pvalue_str = sprintf("%.3f", pvalue_raw),
    adj_pvalue_str = sprintf("%.3f", pvalue_adj),
    sig_star = ifelse(pvalue_raw < 0.05, "*", ""),
    feature_label = paste0(feature_clean, sig_star),
    sample_sizes = sapply(base_data$feature, function(f) {sum(!is.na(combined[[f]]))}),
    group0 = paste(group0,"/", sample_sizes, sep = ""),
    group1 = paste(group1,"/", sample_sizes, sep = "")

  )

# Create label text
labeltext <- rbind(
  c("Feature", "Negative", "Positive", "OR [95% CI]", "P-Value", "Adj. P-Value"),
  as.matrix(cbind(
    base_data$feature_label,
    base_data$group0,
    base_data$group1,
    base_data$estimate_ci,
    base_data$pvalue_str,
    base_data$adj_pvalue_str  ))
)

# Add NA to align rows with label header
mean_all  <- c(NA, base_data$mean)
lower_all <- c(NA, base_data$lower)
upper_all <- c(NA, base_data$upper)

png("forest_plot_steroid_exposure.png", width = 1200, height = 650, res = 110)

# Forest plot
forestplot(
  labeltext = labeltext,
  mean = mean_all,
  lower = lower_all,
  upper = upper_all,
  is.summary = c(TRUE, rep(FALSE, nrow(base_data))),
  boxsize = 0.2,
  col = fpColors(box = "slateblue4", line = "mediumorchid4", zero = "red"),
  xlog = TRUE,  # log scale for odds ratios
  clip = c(min(base_data$lower, na.rm = TRUE), max(base_data$upper, na.rm = TRUE)),
  xticks = pretty(c(base_data$lower, base_data$upper)),
  title = "Impact of Steroid Exposure on Vascular Histopathology",
  txt_gp = fpTxtGp(
    label = gpar(cex = 0.8),
    ticks = gpar(cex = 0.8),
    xlab = gpar(cex = 0.9)
  ))

grid.text("*P-values adjusted with Benjamini-Hochberg (FDR) procedure", x = 0.01, y = unit(0.1, "npc") - unit(2, "lines"), 
          just = "left", gp = gpar(cex = 0.8, col = "black"))
dev.off()

```

Sex Plots
```{r}
sex_base_data <- tibble(
  mean  = sex_summary$OR,
  lower = sex_summary$lowerCI,
  upper = sex_summary$upperCI,
  feature = sex_summary$feature,
  pvalue_raw = sex_summary$pvalue,
  pvalue_adj = sex_summary$AdjP,
  estimate_ci = paste0(
    sprintf("%.2f", sex_summary$OR), " [",
    sprintf("%.2f", sex_summary$lowerCI), ", ",
    sprintf("%.2f", sex_summary$upperCI), "]"
  ),
  group0 = steroid_summary$group0,
  group1 = steroid_summary$group1
) %>%
  mutate(
    feature_clean = str_replace_all(feature, "_", " "),
    feature_clean = str_remove(feature_clean, " ?bin$"),
    feature_clean = str_to_title(feature_clean),
    pvalue_str = sprintf("%.3f", pvalue_raw),
    adj_pvalue_str = sprintf("%.3f", pvalue_adj),
    sig_star = ifelse(pvalue_raw < 0.05, "*", ""),
    feature_label = paste0(feature_clean, sig_star),
    sample_sizes = sapply(sex_summary$feature, function(f) {sum(!is.na(combined[[f]]))}),
    group0 = paste(group0,"/", sample_sizes, sep = ""),
    group1 = paste(group1,"/", sample_sizes, sep = "")

  )


# Filter NA or problematic values
sex_base_data <- sex_base_data %>%
  filter(!is.na(mean) & !is.na(lower) & !is.na(upper))

# Label matrix
sex_labeltext <- rbind(
  c("Feature", "Negative", "Positive", "OR [95% CI]", "P-Value", "Adj. P-Value"),
  as.matrix(cbind(
    sex_base_data$feature_label,
    base_data$group0,
    base_data$group1,
    sex_base_data$estimate_ci,
    sex_base_data$pvalue_str,
    sex_base_data$adj_pvalue_str
  ))
)

# Prepare values
mean_all  <- c(NA, sex_base_data$mean)
lower_all <- c(NA, sex_base_data$lower)
upper_all <- c(NA, sex_base_data$upper)

png("forest_plot_sex.png", width = 1200, height = 650, res = 110)

# Plot on linear scale
forestplot(
  labeltext = sex_labeltext,
  mean = mean_all,
  lower = lower_all,
  upper = upper_all,
  is.summary = c(TRUE, rep(FALSE, nrow(sex_base_data))),
  boxsize = 0.2,
  col = fpColors(box =  "royalblue", line = "darkblue", zero = "red"),
  xlog = FALSE,  # linear scale for log-odds
  clip = c(min(lower_all, na.rm = TRUE), max(upper_all, na.rm = TRUE)),
  xticks = pretty(c(min(lower_all, na.rm = TRUE), max(upper_all, na.rm = TRUE))),
  title = "Influence of Patient Sex on Histological Features",
  txt_gp = fpTxtGp(
    label = gpar(cex = 0.8),
    ticks = gpar(cex = 0.8),
    xlab = gpar(cex = 0.9)
  )
)

grid.text("*P-values adjusted with Benjamini-Hochberg (FDR) procedure", x = 0.01, y = unit(0.1, "npc") - unit(2, "lines"), 
          just = "left", gp = gpar(cex = 0.8, col = "black"))

dev.off()


```


Age Plots
```{r}
# Prepare the data for forestplot
age_base_data <- tibble(
  mean  = age_summary$OR,
  lower = age_summary$LowerCI,
  upper = age_summary$UpperCI,
  feature = age_summary$Feature,
  pvalue_raw = age_summary$PValue,
  pvalue_adj = age_summary$AdjP,
  estimate_ci = paste0(
    sprintf("%.2f", age_summary$OR), " [",
    sprintf("%.2f", age_summary$LowerCI), ", ",
    sprintf("%.2f", age_summary$UpperCI), "]"
  ),
  group0 = steroid_summary$group0,
  group1 = steroid_summary$group1
) %>%
  mutate(
    # Clean and label features
    feature_clean = str_replace_all(feature, "_", " "),
    feature_clean = str_remove(feature_clean, " ?bin$"),
    feature_clean = str_to_title(feature_clean),

    # Format p-values
    pvalue_str = sprintf("%.3f", pvalue_raw),
    adj_pvalue_str = sprintf("%.3f", pvalue_adj),
    sig_star = ifelse(pvalue_raw < 0.05, "*", ""),
    feature_label = paste0(feature_clean, sig_star),
    sample_sizes = sapply(age_base_data$feature, function(f) {sum(!is.na(combined[[f]]))}),
    group0 = paste(group0,"/", sample_sizes, sep = ""),
    group1 = paste(group1,"/", sample_sizes, sep = "")
  )


# Prepare the label text for the forest plot
age_labeltext <- rbind(
  c("Feature", "Negative", "Positive", "OR [95% CI]", "P-Value", "Adj. P-Value"),
  as.matrix(cbind(
    age_base_data$feature_label,
    base_data$group0,
    base_data$group1,
    age_base_data$estimate_ci,
    age_base_data$pvalue_str,
    age_base_data$adj_pvalue_str
    ))
)


# Add NA to align with the header row
mean_all  <- c(NA, age_base_data$mean)
lower_all <- c(NA, age_base_data$lower)
upper_all <- c(NA, age_base_data$upper)

# Safe clipping range
min_val <- max(min(age_base_data$lower, na.rm = TRUE), 0.01)
max_val <- max(age_base_data$upper, na.rm = TRUE)

png("forest_plot_age.png", width = 1200, height = 650, res = 110)

# Forest plot
forestplot(
  labeltext = age_labeltext,
  mean = mean_all,
  lower = lower_all,
  upper = upper_all,
  is.summary = c(TRUE, rep(FALSE, nrow(age_base_data))),
  boxsize = 0.2,
  col = fpColors(box = "darkgreen", line = "forestgreen", zero = "red"),
  xlog = TRUE,
  clip = c(min_val, max_val),
  xticks = pretty(c(min_val, max_val)),
  title = "Impact of Age on Vascular Histopathology",
  txt_gp = fpTxtGp(
    label = gpar(cex = 0.8),
    ticks = gpar(cex = 0.8),
    xlab  = gpar(cex = 0.9)
  )
)

grid.text("*P-values adjusted with Benjamini-Hochberg (FDR) procedure", x = 0.01, y = unit(0.1, "npc") - unit(2, "lines"), 
          just = "left", gp = gpar(cex = 0.8, col = "black"))

dev.off()
```


