---
title: "Covariate Analysis"
output: html_document
---

Load relevant data. 
```{r,  include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(ggplot2)
library(dplyr)
library(MASS)
library(scales)
library(broom)
library(gridExtra)
library(tibble)
library(stringr)
library(forestplot)
library(grid)

clin_var <- read.csv('/Users/kristinemedcalf/Desktop/Programming/Dissertation_GCA_Prediction/data/clinical_variables_cc_c1_c2.csv')
hist_var <- read.csv("/Users/kristinemedcalf/Desktop/Programming/Dissertation_GCA_Prediction/data/histological_variables_cc_c1_c2.csv")
names(hist_var)[names(hist_var) == "Lymphocytic_infiltrate_in._adventitia"] <- "Lymphocytic_infiltrate_in_adventitia"

combined <- merge(clin_var, hist_var, by = "ID")

```

Remove features with high correlation, low variance, or low usefulness: 
• Any_granulomatous_infiltrate
• PALI
• Any_lymphocytic_infiltrate
• GCA_present
• Granulomatous_infiltrate_in_media
• Hyperplasia
• Fibrosis

```{r}
cols_to_remove <- c("Any_granulomatous_infiltrate", "PALI", "Any_lymphocytic_infiltrate", "GCA_present", "Granulomatous_infiltrate_in_media", "Hyperplasia", "Fibrosis")
hist_var <- hist_var[, setdiff(names(hist_var), cols_to_remove)]
trimmed_hist <- as.data.frame(hist_var[, -1])
```

MODEL TYPES:
Logistic Models: continuous or categorical features, binary outcomes
Linear Models: continuous (or encoded categorical) features, continuous outcomes
Classification Models: continuous or categorical features, categorical outcomes 

Variables:
Steroids_days → continuous
Sex → binary
Age (quartiles) → categorical (ordinal)

Modeling each outcome:
Binary features → Steroids_days: Linear model
Binary features → Sex: Logistic model
Binary features → Age (quartiles): Classification model

Group different feature types (binary or categorical)
```{r}
trimmed_hist <- hist_var[, -1]  

# retrieve binary
binary_features <- c()
for (col in names(trimmed_hist)) {
  if (length(levels(factor(trimmed_hist[[col]]))) == 2) {
    binary_features <- c(binary_features, col)
  }
}

# retrieve categorical
categorical_features <- setdiff(names(trimmed_hist), binary_features)

for (col in c(binary_features, categorical_features)) {
  combined[[col]] <- factor(combined[[col]])
}
all_features <- c(binary_features, categorical_features)

```

Steroid Exposure Models
```{r}
# create a linear model for each feature and store in df
steroid_model <- lapply(all_features, function(feature) {
  lm(as.formula(paste("steroids_days ~", feature)), data = combined)
})

# name each linear model (row) in df according to feature
names(steroid_model) <- all_features

# retrieve all relevant data for each steroid model
steroid_summary <- lapply(seq_along(steroid_model), function(i) {
  model <- steroid_model[[i]]
  feature <- all_features[i]
  coef_summary <- summary(model)$coefficients
  coef_summary <- coef_summary[rownames(coef_summary) != "(Intercept)", , drop = FALSE]

  data.frame(
    feature = if (feature %in% categorical_features) rownames(coef_summary) else feature,
    estimate = coef_summary[, "Estimate"],
    lowerCI = coef_summary[, "Estimate"] - 1.96 * coef_summary[, "Std. Error"],
    upperCI = coef_summary[, "Estimate"] + 1.96 * coef_summary[, "Std. Error"],
    pvalue = coef_summary[, "Pr(>|t|)"]
  )
})

steroid_summary <- do.call(rbind, steroid_summary)
steroid_summary$AdjP <- p.adjust(steroid_summary$pvalue, method = "fdr")
```

Sex Models
```{r}
# change sex into binary inputs
combined$sex <- factor(combined$sex, levels = c(1, 2), labels = c("Male", "Female"))

# create a logistic model for each feature and store in df
sex_model <- lapply(all_features, function(feature) {
  glm(as.formula(paste("sex ~", feature)), data = combined, family = binomial)
})

names(sex_model) <- all_features

# retrieve all relevant data for each sex model
sex_summary <- lapply(seq_along(sex_model), function(i) {
  model <- sex_model[[i]]
  feature <- all_features[i]
  coef_summary <- summary(model)$coefficients
  coef_summary <- coef_summary[rownames(coef_summary) != "(Intercept)", , drop = FALSE]

  data.frame(
    feature = if (feature %in% categorical_features) rownames(coef_summary) else feature,
    estimate = exp(coef_summary[, "Estimate"]),
    lowerCI = exp(coef_summary[, "Estimate"] - 1.96 * coef_summary[, "Std. Error"]),
    upperCI = exp(coef_summary[, "Estimate"] + 1.96 * coef_summary[, "Std. Error"]),
    pvalue = coef_summary[, "Pr(>|z|)"]
  )
})

sex_summary <- do.call(rbind, sex_summary)
sex_summary$AdjP <- p.adjust(sex_summary$pvalue, method = "fdr")
```

Age Models
```{r}
# ensure 1st level is used as baseline
combined$Media_pattern      <- factor(combined$Media_pattern, levels = c("1", "2", "3"))
combined$Intima_pattern     <- factor(combined$Intima_pattern, levels = c("1", "2", "3"))
combined$Adventitia_pattern <- factor(combined$Adventitia_pattern, levels = c("1", "2", "3"))
combined$Barcelona_score    <- factor(combined$Barcelona_score, levels = c("1", "2", "3", "4"))
combined$Occlusion_grade    <- factor(combined$Occlusion_grade, levels = c("1", "2", "3", "4"))

# group age into groups of 11 years
min_age <- floor(min(combined$age, na.rm = TRUE))
max_age <- ceiling(max(combined$age, na.rm = TRUE))
eleven_year_breaks <- seq(min_age, max_age + 11, by = 11)

combined$age_with_eleven_year_breaks <- cut(
  combined$age,
  breaks = eleven_year_breaks,
  include.lowest = TRUE,
  right = FALSE
)

levels(combined$age_with_eleven_year_breaks) <- c("<70", "70–80", ">80")

# Fit model to data <- perform three glm (one for each age group) 
# <70 used as baseline
extract_glm_summary <- function(data, feature, level_label) {
  formula <- as.formula(paste("binary_age ~", feature))
  model <- glm(formula, data = data, family = binomial)

  coefs <- summary(model)$coefficients
  coefs <- coefs[rownames(coefs) != "(Intercept)", , drop = FALSE]

  if (nrow(coefs) == 0) {
    return(NULL)
  }
  
  # retrieve model results and save to df
  results <- lapply(seq_len(nrow(coefs)), function(i) {
    row <- coefs[i, ]
    est <- row["Estimate"]
    se <- row["Std. Error"]
    p <- row["Pr(>|z|)"]

    data.frame(
      Feature = feature,
      Level = rownames(coefs)[i],
      Contrast = level_label,
      OR = exp(est),
      LowerCI = exp(est - 1.96 * se),
      UpperCI = exp(est + 1.96 * se),
      PValue = p
    )
  })

  do.call(rbind, results)
}

# combine all results
results <- list()

for (feature in all_features) {
  # 70–80 vs <70
  df_mid <- combined %>%
    filter(age_with_eleven_year_breaks %in% c("<70", "70–80")) %>%
    mutate(binary_age = factor(age_with_eleven_year_breaks, levels = c("<70", "70–80")))
  res_mid <- extract_glm_summary(df_mid, feature, "70–80 vs <70")

  # >80 vs <70
  df_old <- combined %>%
    filter(age_with_eleven_year_breaks %in% c("<70", ">80")) %>%
    mutate(binary_age = factor(age_with_eleven_year_breaks, levels = c("<70", ">80")))
  res_old <- extract_glm_summary(df_old, feature, ">80 vs <70")

  results[[feature]] <- bind_rows(res_mid, res_old)
}

age_binary_summary <- bind_rows(results)
age_binary_summary$AdjP <- p.adjust(age_binary_summary$PValue, method = "fdr")

```


Forest Plots

Steroid Exposure Plots
```{r}
# change labels for categorical features
occlusion_labels <- c("1" = "Minimal Occlusion", "2" = "Mild Occlusion",
                      "3" = "Moderate Occlusion", "4" = "Severe Occlusion")

barcelona_labels <- c("1" = "Grade I", "2" = "Grade II",
                      "3" = "Grade III", "4" = "Grade IV")

pattern_labels <- c("1" = "Regular", "2" = "Irregular", "3" = "Disrupted")

feature_labels <- c(occlusion_labels, barcelona_labels, pattern_labels)

base_data <- tibble(
  mean  = steroid_summary$estimate,
  lower = steroid_summary$lowerCI,
  upper = steroid_summary$upperCI,
  feature = steroid_summary$feature,
  pvalue = steroid_summary$pvalue,
  estimate_ci = paste0(
    sprintf("%.2f", steroid_summary$estimate), " [",
    sprintf("%.2f", steroid_summary$lowerCI), ", ",
    sprintf("%.2f", steroid_summary$upperCI), "]"
  )
) %>%
  mutate(
    feature_clean = str_trim(str_to_lower(feature)),
    feature_clean = str_replace_all(feature_clean, "_", " "),
    feature_clean = case_when(
      str_detect(feature_clean, "occlusion grade\\d") ~ paste0(
        "Occlusion Grade: ", occlusion_labels[str_extract(feature_clean, "\\d+$")]
      ),
      str_detect(feature_clean, "barcelona score\\d") ~ paste0(
        "Barcelona Score: ", barcelona_labels[str_extract(feature_clean, "\\d+$")]
      ),
      str_detect(feature_clean, "intima pattern\\d") ~ paste0(
        "Intima Pattern: ", pattern_labels[str_extract(feature_clean, "\\d+$")]
      ),
      str_detect(feature_clean, "media pattern\\d") ~ paste0(
        "Media Pattern: ", pattern_labels[str_extract(feature_clean, "\\d+$")]
      ),
      str_detect(feature_clean, "adventitia pattern\\d") ~ paste0(
        "Adventitia Pattern: ", pattern_labels[str_extract(feature_clean, "\\d+$")]
      ),
      TRUE ~ str_to_title(feature_clean)
    ),
    pvalue_str = sprintf("%.3f", pvalue),
    sig_star = ifelse(pvalue < 0.05, "*", ""),
    feature_label = paste0(feature_clean, sig_star)
  )

data1 <- base_data[1:14, ]
data2 <- base_data[15:28, ]

labeltext1 <- rbind(
  c("Feature", "Estimate (95% CI)", "P value"),
  as.matrix(cbind(data1$feature_label, data1$estimate_ci, data1$pvalue_str))
)

labeltext2 <- rbind(
  c("Feature", "Estimate (95% CI)", "P value"),
  as.matrix(cbind(data2$feature_label, data2$estimate_ci, data2$pvalue_str))
)

mean1 <- c(NA, data1$mean)
lower1 <- c(NA, data1$lower)
upper1 <- c(NA, data1$upper)

mean2 <- c(NA, data2$mean)
lower2 <- c(NA, data2$lower)
upper2 <- c(NA, data2$upper)

# plot 1
forestplot(
  labeltext = labeltext1,
  mean = mean1,
  lower = lower1,
  upper = upper1,
  is.summary = c(TRUE, rep(FALSE, nrow(data1))),
  boxsize = 0.2,
  col = fpColors(box = "royalblue", line = "darkblue", zero = "red"),
  xlog = FALSE,
  clip = c(min(base_data$lower, na.rm = TRUE), max(base_data$upper, na.rm = TRUE)),
  xticks = pretty(c(base_data$lower, base_data$upper)),
  title = "Steroid Exposure Predictors (1 of 2)",
  txt_gp = fpTxtGp(
    label = gpar(cex = 0.8),
    ticks = gpar(cex = 0.8),
    xlab = gpar(cex = 0.9)
  )
)

# plot 2
forestplot(
  labeltext = labeltext2,
  mean = mean2,
  lower = lower2,
  upper = upper2,
  is.summary = c(TRUE, rep(FALSE, nrow(data2))),
  boxsize = 0.2,
  col = fpColors(box = "royalblue", line = "darkblue", zero = "red"),
  xlog = FALSE,
  clip = c(min(base_data$lower, na.rm = TRUE), max(base_data$upper, na.rm = TRUE)),
  xticks = pretty(c(base_data$lower, base_data$upper)),
  title = "Steroid Exposure Predictors (2 of 2)",
  txt_gp = fpTxtGp(
    label = gpar(cex = 0.8),
    ticks = gpar(cex = 0.8),
    xlab = gpar(cex = 0.9)
  )
)

```
Sex Plots
```{r}
base_data <- tibble(
  mean  = sex_summary$estimate,
  lower = sex_summary$lowerCI,
  upper = sex_summary$upperCI,
  feature = sex_summary$feature,
  pvalue = sex_summary$pvalue,
  estimate_ci = paste0(
    sprintf("%.2f", sex_summary$estimate), " [",
    sprintf("%.2f", sex_summary$lowerCI), ", ",
    sprintf("%.2f", sex_summary$upperCI), "]"
  )
) %>%
  mutate(
    feature_clean = str_trim(str_to_lower(feature)),
    feature_clean = str_replace_all(feature_clean, "_", " "),
    feature_clean = case_when(
      str_detect(feature_clean, "occlusion grade\\d") ~ paste0(
        "Occlusion Grade: ", occlusion_labels[str_extract(feature_clean, "\\d+$")]
      ),
      str_detect(feature_clean, "barcelona score\\d") ~ paste0(
        "Barcelona Score: ", barcelona_labels[str_extract(feature_clean, "\\d+$")]
      ),
      str_detect(feature_clean, "intima pattern\\d") ~ paste0(
        "Intima Pattern: ", pattern_labels[str_extract(feature_clean, "\\d+$")]
      ),
      str_detect(feature_clean, "media pattern\\d") ~ paste0(
        "Media Pattern: ", pattern_labels[str_extract(feature_clean, "\\d+$")]
      ),
      str_detect(feature_clean, "adventitia pattern\\d") ~ paste0(
        "Adventitia Pattern: ", pattern_labels[str_extract(feature_clean, "\\d+$")]
      ),
      TRUE ~ str_to_title(feature_clean)
    ),
    pvalue_str = sprintf("%.3f", pvalue),
    sig_star = ifelse(pvalue < 0.05, "*", ""),
    feature_label = paste0(feature_clean, sig_star)
  )

data1 <- base_data[1:14, ]
data2 <- base_data[15:nrow(base_data), ]

labeltext1 <- rbind(
  c("Feature", "Estimate (95% CI)", "P Value"),
  as.matrix(cbind(data1$feature_label, data1$estimate_ci, data1$pvalue_str))
)

labeltext2 <- rbind(
  c("Feature", "Estimate (95% CI)", "P Value"),
  as.matrix(cbind(data2$feature_label, data2$estimate_ci, data2$pvalue_str))
)

mean1 <- c(NA, data1$mean)
lower1 <- c(NA, data1$lower)
upper1 <- c(NA, data1$upper)

mean2 <- c(NA, data2$mean)
lower2 <- c(NA, data2$lower)
upper2 <- c(NA, data2$upper)

# plot 1
forestplot(
  labeltext = labeltext1,
  mean = mean1,
  lower = lower1,
  upper = upper1,
  is.summary = c(TRUE, rep(FALSE, nrow(data1))),
  xlog = TRUE,
  boxsize = 0.2,
  col = fpColors(box = "darkblue", line = "blue", zero = "red"),
  xticks = c(0.5, 1, 2, 4),
  clip = c(0.1, 5),
  title = "Histological Predictors of Patient Sex (1 of 2)",
  txt_gp = fpTxtGp(
    label = gpar(cex = 0.8),
    ticks = gpar(cex = 0.8),
    xlab  = gpar(cex = 0.9)
  )
)

# plot 2
forestplot(
  labeltext = labeltext2,
  mean = mean2,
  lower = lower2,
  upper = upper2,
  is.summary = c(TRUE, rep(FALSE, nrow(data2))),
  xlog = TRUE,
  boxsize = 0.2,
  col = fpColors(box = "darkblue", line = "blue", zero = "red"),
  xticks = c(0.5, 1, 2, 4),
  clip = c(0.1, 5),
  title = "Histological Predictors of Patient Sex (2 of 2)",
  txt_gp = fpTxtGp(
    label = gpar(cex = 0.8),
    ticks = gpar(cex = 0.8),
    xlab  = gpar(cex = 0.9)
  )
)

```


Age Plots
```{r}

plot_data <- age_binary_summary %>%
  mutate(
    is_infinite = !is.finite(UpperCI) | !is.finite(LowerCI) | !is.finite(OR),
    OR_CI = ifelse(
      is_infinite,
      "Unstable",
      paste0(sprintf("%.2f", OR), " [", sprintf("%.2f", LowerCI), ", ", sprintf("%.2f", UpperCI), "]")
    ),
    
    raw_feature = str_extract(Feature, "^[^:]+"),
    feature_clean = str_to_lower(raw_feature),
    feature_clean = str_replace_all(feature_clean, "[_:]", " "),
    feature_clean = str_squish(feature_clean),
    
    feature_clean = case_when(
      str_detect(feature_clean, "occlusion grade") ~ paste0(
        "Occlusion Grade: ", occlusion_labels[str_extract(Level, "\\d+$")]
      ),
      str_detect(feature_clean, "barcelona score") ~ paste0(
        "Barcelona Score: ", barcelona_labels[str_extract(Level, "\\d+$")]
      ),
      str_detect(feature_clean, "intima pattern") ~ paste0(
        "Intima Pattern: ", pattern_labels[str_extract(Level, "\\d+$")]
      ),
      str_detect(feature_clean, "media pattern") ~ paste0(
        "Media Pattern: ", pattern_labels[str_extract(Level, "\\d+$")]
      ),
      str_detect(feature_clean, "adventitia pattern") ~ paste0(
        "Adventitia Pattern: ", pattern_labels[str_extract(Level, "\\d+$")]
      ),
      TRUE ~ str_to_title(feature_clean)
    ),
    
    Contrast = factor(Contrast, levels = c("70–80 vs <70", ">80 vs <70")),
    Label = paste0(feature_clean, " (", Contrast, ")"),
    SigStar = ifelse(PValue < 0.05, "*", ""),
    LabelFull = paste0(Label, SigStar),
    Pval_str = sprintf("%.3f", PValue)
  ) %>%
  arrange(Contrast, feature_clean)


n <- nrow(plot_data)
chunk_size <- ceiling(n / 4)

data1 <- plot_data[1:min(chunk_size, n), ]
data2 <- plot_data[(chunk_size + 1):min(2 * chunk_size, n), ]
data3 <- plot_data[(2 * chunk_size + 1):min(3 * chunk_size, n), ]
data4 <- plot_data[(3 * chunk_size + 1):n, ]

make_forest_plot <- function(data_chunk, title_text) {
  if (nrow(data_chunk) == 0) return(NULL)

  labeltext <- rbind(
    c("Feature", "OR (95% CI)", "P Value"),
    as.matrix(cbind(data_chunk$LabelFull, data_chunk$OR_CI, data_chunk$Pval_str))
  )

  forestplot(
    labeltext = labeltext,
    mean = c(NA, data_chunk$OR),
    lower = c(NA, data_chunk$LowerCI),
    upper = c(NA, data_chunk$UpperCI),
    is.summary = c(TRUE, rep(FALSE, nrow(data_chunk))),
    xlog = TRUE,
    boxsize = 0.25,
    xticks = c(0.5, 1, 2, 4),
    clip = c(0.1, 10),
    title = title_text,
    txt_gp = fpTxtGp(
      label = gpar(cex = 0.8),
      ticks = gpar(cex = 0.8),
      xlab  = gpar(cex = 0.9)
    ),
    col = fpColors(box = "navy", line = "blue", zero = "red")
  )
}

make_forest_plot(data1, "Age Group Predictors (Plot 1 of 4)")
make_forest_plot(data2, "Age Group Predictors (Plot 2 of 4)")
make_forest_plot(data3, "Age Group Predictors (Plot 3 of 4)")
make_forest_plot(data4, "Age Group Predictors (Plot 4 of 4)")


```
* level 1s are baselines and are therefore excluded from model summaries.

