---
title: "Disseration Summary"
author: "Kristine Medcalf"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
  pdf_document:
    toc: true
    toc_depth: 2

---

## 1. Introduction
___


## 2. Data Preparation
___
Two datasets were used:

- **Clinical Variables**: contains variables regarding the individual such as age, sex, and steriod exposure (in days)
- **Histological Variables**: contains variables of symptoms and other variables related to the disease.

```{r, echo=FALSE, message = FALSE, warning=FALSE}
library(tidyr)
library(knitr)
library(ggplot2)
library(dplyr)
library(MASS) 
library(scales)
library(broom)
library(gridExtra)
library(tibble)
library(stringr)
library(kableExtra)
library(DESeq2)
library(forestplot)
library(biomaRt)
library(grid)

clin_var <-  read.csv('/Users/kristinemedcalf/Desktop/Programming/Dissertation_GCA_Prediction/data/clinical_variables_cc_c1_c2.csv')
hist_var <- read.csv("/Users/kristinemedcalf/Desktop/Programming/Dissertation_GCA_Prediction/data/histological_variables_cc_c1_c2.csv")
names(hist_var)[names(hist_var) == "Lymphocytic_infiltrate_in._adventitia"] <- "Lymphocytic_infiltrate_in_adventitia"
```

<details><summary><strong>View clinical variables</strong></summary>

```{r, results='asis', echo=FALSE}
vars <- names(clin_var)[-1]
cat(paste0("- ", vars, collapse = "\n"))

```

</details>

<pr>

<details><summary><strong>View histological variables</strong></summary>

```{r, results='asis', echo=FALSE}
vars <- names(hist_var)[-1]
cat(paste0("- ", vars, collapse = "\n"))
cols_to_remove <- c("Any_granulomatous_infiltrate", "PALI", "Any_lymphocytic_infiltrate", "GCA_present", 'Barcelona_score')
hist_var <- hist_var[, setdiff(names(hist_var), cols_to_remove)]
```
</details>

<pr>

The data cleaning process focuses on identifying **redundant** and **collinear** variables within the histological dataset. Features with minimal variability or near-constant distributions were also removed, as they contributed little to the modeling process. These steps helped:

- Reduce *noise* and *overfitting*
- Improve *model interpretability* and *stability*
- Lower the *risk of false positives* in downstream analyses

Overall, this refinement ensured that each feature retained for analysis adds distinct and meaningful information.

Specifically, the following variables were removed due to low variance, redundancy, or lack of added value:

- **Any_granulomatous_infiltrate,**
- **PALI,**
- **Any_lymphocytic_infiltrate,**
- **GCA_present,** and 
- **Barcelona_score.**

<pr>

### 2.1 Exploratory Cross-Tabulations
___

<pr>

#### 2.1.1 Stromal Features vs Occlusion Grade

To assess redundancy among stromal response features, cross-tabulations were performed between Hyperplasia, Oedema, Fibrosis, and Occlusion Grade. These help determine whether a single variable (Occlusion Grade) can summarise similar information.

```{r, echo=FALSE}
tab1 <- as.data.frame.matrix(xtabs(~ Hyperplasia + Occlusion_grade, data = hist_var))
tab1 <- rownames_to_column(tab1, var = "Hyperplasia")
kable(tab1, caption = " ", row.names = FALSE) %>%
  kable_styling(full_width = FALSE, position = "left", bootstrap_options = c("striped", "hover")) %>%
  add_header_above(c(" " = 1, "Occlusion Grade" = ncol(tab1) - 1))
```

```{r, echo=FALSE}
tab1 <- as.data.frame.matrix(xtabs(~ Oedema + Occlusion_grade, data = hist_var))
tab1 <- rownames_to_column(tab1, var = "Oedema")
kable(tab1, caption = " ", row.names = FALSE) %>%
  kable_styling(full_width = FALSE, position = "left", bootstrap_options = c("striped", "hover")) %>%
  add_header_above(c(" " = 1, "Occlusion Grade" = ncol(tab1) - 1))
```

```{r, echo=FALSE}
tab1 <- as.data.frame.matrix(xtabs(~ Fibrosis + Occlusion_grade, data = hist_var))
tab1 <- rownames_to_column(tab1, var = "Fibrosis")
kable(tab1, caption = " ", row.names = FALSE) %>%
  kable_styling(full_width = FALSE, position = "left", bootstrap_options = c("striped", "hover")) %>%
  add_header_above(c(" " = 1, "Occlusion Grade" = ncol(tab1) - 1))
```

These cross-tabulations demonstrate that Hyperplasia and Fibrosis are strongly associated with higher Occlusion Grades. Nearly all cases exhibiting Hyperplasia fall within Occlusion Grades 3 or 4, and Fibrosis follows a similar distribution pattern. This suggests that these features may be reflecting the same underlying stromal response.

Given this overlap, it may be appropriate to retain Occlusion Grade as the primary stromal variable and exclude Hyperplasia and Fibrosis from further analysis. In contrast, Oedema shows a slightly more varied distribution across grades, which may indicate it captures an additional or distinct aspect of stromal change, and could therefore be retained.
<pr>

#### 2.1.2 Immune Cell Infiltrates vs Tissue Layer Patterns

Cross-tabulations were also performed between lymphocytic infiltrate indicators and histological patterns in the adventitia, media, and intima layers. This helps identify whether these represent overlapping immune activity.

```{r, echo=FALSE}
tab1 <- as.data.frame.matrix(xtabs(~ Lymphocytic_infiltrate_in_adventitia + Adventitia_pattern, data = hist_var))
tab1 <- rownames_to_column(tab1, var = "Lymphocytic_infiltrate_in_adventitia")
kable(tab1, caption = " ", row.names = FALSE) %>%
  kable_styling(full_width = FALSE, position = "left", bootstrap_options = c("striped", "hover")) %>%
  add_header_above(c(" " = 1, "Adventitia_pattern" = ncol(tab1) - 1))
```

```{r, echo=FALSE}
tab1 <- as.data.frame.matrix(xtabs(~ Lymphocytic_infiltrate_in_media + Media_pattern, data = hist_var))
tab1 <- rownames_to_column(tab1, var = "Lymphocytic_infiltrate_in_media")
kable(tab1, caption = " ", row.names = FALSE) %>%
  kable_styling(full_width = FALSE, position = "left", bootstrap_options = c("striped", "hover")) %>%
  add_header_above(c(" " = 1, "Media_pattern" = ncol(tab1) - 1))
```

```{r, echo=FALSE}
tab1 <- as.data.frame.matrix(xtabs(~ Lymphocytic_infiltrate_in_intima + Intima_pattern, data = hist_var))
tab1 <- rownames_to_column(tab1, var = "Lymphocytic_infiltrate_in_intima")
kable(tab1, caption = " ", row.names = FALSE) %>%
  kable_styling(full_width = FALSE, position = "left", bootstrap_options = c("striped", "hover")) %>%
  add_header_above(c(" " = 1, "Intima_pattern" = ncol(tab1) - 1))
```

The cross-tabulations reveal a strong alignment between lymphocytic infiltrates and their respective vessel wall patterns, specifically, the adventitia, media, and intima layers. For example, cases with lymphocytic infiltrate in the adventitia consistently correspond to higher values in the Adventitia Pattern variable. Similar trends are observed for the media and intima.

This suggests that the pattern variables may already be capturing the presence and extent of lymphocytic infiltration within each vessel layer. Therefore, it may be appropriate to retain only the Adventitia/Media/Intima Pattern variables in downstream analyses, and drop the direct infiltrate indicators, as they likely do not add independent information.

```{r, echo = FALSE}
# Define columns to remove based on redundancy
cols_to_remove <- c(
  "Hyperplasia",
  "Fibrosis",
  "Lymphocytic_infiltrate_in_adventitia",
  "Lymphocytic_infiltrate_in_media",
  "Lymphocytic_infiltrate_in_intima"
)

# Remove them from the dataframe (if they exist)
hist_var <- hist_var[, !(names(hist_var) %in% cols_to_remove)]
```

<pr>

### 2.2 Variance Analysis
<pr>

#### 2.2.1 Frequency Tables

To understand the data, we first summarized **histological features** by showing the frequency of each category.

Only features with a small number of unique numeric values were treated as factors. This approach allowed a clearer summary of the data.

<details><summary><strong>View summary table of feature distributions</strong></summary>

*This includes the count of samples in each category across multiple histological variables.*

```{r, echo = FALSE}
trimmed_hist <- as.data.frame(hist_var[, -1])

for (col in names(trimmed_hist)) {
  if (is.numeric(trimmed_hist[[col]]) && length(unique(trimmed_hist[[col]])) < 10) {
    trimmed_hist[[col]] <- as.factor(trimmed_hist[[col]])
  }
}

grouped_cols <- sapply(trimmed_hist, function(x) is.factor(x) || is.character(x))
grouped_features <- names(trimmed_hist)[grouped_cols]

grouped_summary <- lapply(grouped_features, function(col) {
  tab <- table(trimmed_hist[[col]])
  df <- as.data.frame(tab)
  names(df) <- c("Level", "Count")
  df$Feature <- col
  return(df)
})

hist_feature_summary <- do.call(rbind, grouped_summary)
wide_hist_summary <- pivot_wider(hist_feature_summary,
                                  names_from = Level,
                                  values_from = Count,
                                  values_fill = NA)

wide_hist_summary <- wide_hist_summary[order(wide_hist_summary$Feature), ]

knitr::kable(wide_hist_summary, caption = "Summary of Histological Feature Frequencies")

```

<pr>

#### 2.2.2 Binary Concordance Calculations

We computed the variance of each histological feature to identify features that do not vary much between samples. **Low-variance features** provide limited discriminative value. A range of features have variances close to 0. 

<details><summary><strong>View histogram of histological feature variance</strong></summary>
```{r, echo = FALSE}
variances <- sapply(trimmed_hist, function(x) {
  if (is.factor(x)) x <- as.numeric(x)
  var(x, na.rm = TRUE)
})

hist(variances, 
     main = "Histogram of Feature Variances",
     xlab = "Variance", 
     ylab = "Frequency",
     col = "lightblue", 
     breaks = 50, 
     xlim = c(0, 2))

```

</details>

<pr>

```{r, echo = FALSE}
low_var_01 <- data.frame(sort(variances[variances < 0.1]))
names(low_var_01) <- c('Variance')

knitr::kable(low_var_01, caption = "Low Variance Histological Features")

```

Since these features exhibit extremely low variance, they are unlikely to contribute meaningful information to the analysis and can therefore be removed.

```{r, echo = FALSE}
cols_to_remove <- c(
  "Aggregates",
  "Granulomatous_infiltrate_in_media",
  "Granulomatous_infiltrate_in_intima",
  "Granulomatous_infiltrate_in_adventitia"
)

hist_var <- hist_var[, !(names(hist_var) %in% cols_to_remove)]
```
<pr>

### 2.3. Correlation Analysis
___

<pr>

#### 2.3.1 Correlation Heatmap

We computed a **correlation matrix** across the remaining numeric histological features and visualized it as a heatmap. Strong correlations between features may indicate redundancy.

```{r, echo=FALSE}
library(pheatmap)
phenotypes <- names(hist_var)[-1]
cor_matrix <- cor(hist_var[, phenotypes], use = "pairwise.complete.obs")
rownames(cor_matrix) <- phenotypes
colnames(cor_matrix) <- phenotypes
pheatmap(cor_matrix,
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         cellheight = 8,
         fontsize_row = 7, 
         fontsize_col = 7)
```

There seems to be an area of high correlation with the different layers of pattern, occlusion_grade, and media_destruction. 

#### 2.3.2 Correlation Calculation

To further assess redundancy, we analyzed **binary features** using the **phi coefficient**, a statistical measure of association between two binary variables.

- A phi coefficient **> 0.5** is considered strong.
- Values **> 0.8** are flagged for possible removal due to redundancy.

```{r, echo=FALSE}
binary_features <- c()
for (col in names(trimmed_hist)) {
  if (length(levels(factor(trimmed_hist[[col]]))) == 2) {
    binary_features <- c(binary_features, col)
  }
}

phi_coefficient <- function(x, y) {
  tab <- table(x, y)
  if (!all(dim(tab) == c(2, 2))) {
    return(NA)
  }
  a <- tab[1,1]; b <- tab[1,2]; c <- tab[2,1]; d <- tab[2,2]
  num <- (a * d) - (b * c)
  denom <- sqrt((a + b) * (c + d) * (a + c) * (b + d))
  if (denom == 0) return(NA)
  return(num / denom)
}

n <- length(binary_features)
binaryfrequencytable <- matrix(NA, nrow = n, ncol = n)
colnames(binaryfrequencytable) <- binary_features
rownames(binaryfrequencytable) <- binary_features
diag(binaryfrequencytable) <- 1

df <- as.data.frame(lapply(trimmed_hist[binary_features], factor))
for (i in 1:(n - 1)) {
  for (j in (i + 1):n) {
    phi <- phi_coefficient(df[[i]], df[[j]])
    binaryfrequencytable[i, j] <- phi
    binaryfrequencytable[j, i] <- phi
  }
}

binaryfrequencytable <- as.data.frame(binaryfrequencytable)
```

<details><summary><strong>View matrix of phi coefficients</strong></summary>

```{r, echo = FALSE}
knitr::kable(binaryfrequencytable, caption = "Binary Frequency Table")
```

*This matrix includes pairwise concordance scores between binary features.*

</details>

<pr>

```{r, echo=FALSE}
get_high_phi_pairs <- function(phi_matrix, threshold = 0.5) {
  cols <- colnames(phi_matrix)
  pairs <- list()
  
  for (i in 1:(length(cols) - 1)) {
    for (j in (i + 1):length(cols)) {
      val <- phi_matrix[i, j]
      if (!is.na(val) && abs(val) > threshold) {
        pairs[[length(pairs) + 1]] <- data.frame(
          Feature1 = cols[i],
          Feature2 = cols[j],
          Phi = val
        )
      }
    }
  }
  do.call(rbind, pairs)
}

high_phi_pairs <- get_high_phi_pairs(binaryfrequencytable, threshold = 0.5)
```
```{r, echo = FALSE}
knitr::kable(high_phi_pairs, caption = "Stong Phi Coeffecient Pairs")
```

While several pairs of binary features have Phi coefficients greater than 0.5, none exceed the redundancy threshold of 0.8. This suggests that although some variables are moderately correlated, there is no strong evidence of multicollinearity. Therefore, no features need to be removed based on Phi correlation alone.

<pr>

## 3. Covariate Analysis
___
To evaluate the potential confounding effects of demographic and treatment variables, we conducted covariate analysis using logistic models with steroid exposure, age, and sex as predictors.

### 3.1 Collapsing Categorical Features

In order to do this, all categorical histological variables were converted to binary format for modeling purposes.

For the three variables reflecting inflammation severity across arterial layers (adventitia, media, intima), values were grouped as follows:

- 0 and 1 were merged into a single category representing low or no inflammation
- 2 and 3 were combined to represent more severe inflammation

For the variable representing the degree of arterial occlusion, categories were grouped into:

- 0 and 1: Minimal or no occlusion
- 2, 3, and 4: Moderate to severe occlusion


```{r, echo=FALSE}

# Recode inflammation severity
hist_var$Adventitia_pattern_bin <- ifelse(hist_var$Adventitia_pattern %in% c(0, 1), 0, 1)
hist_var$Media_pattern_bin      <- ifelse(hist_var$Media_pattern %in% c(0, 1), 0, 1)
hist_var$Intima_pattern_bin     <- ifelse(hist_var$Intima_pattern %in% c(0, 1), 0, 1)

hist_var$Occlusion_Grade_bin <- ifelse(
  is.na(hist_var$Occlusion_grade), NA,
  ifelse(hist_var$Occlusion_grade %in% c(0, 1), 0, 1)
)


all_features <- c()
for (col in names(hist_var)) {
  if (length(levels(factor(hist_var[[col]]))) == 2) {
    all_features <- c(all_features, col)
  }
}

combined <- merge(clin_var, hist_var, by = "ID")
hist_var <- as.data.frame(hist_var[, -1])

```

### 3.2 Regression Models

#### 3.2.1 Steroid Exposure Models
___
We examined the association between steroid treatment duration (measured in days) and each binary histological feature using logistic regression. 

```{r, echo=FALSE, warning=FALSE}
# Fit logistic model for each binary feature with steroid_days as predictor
steroid_model <- lapply(all_features, function(feature) {
  glm(as.formula(paste(feature, "~ steroids_days")), 
      family = binomial(link = "logit"), 
      data = combined)
})

# Name each model
names(steroid_model) <- all_features

# Summarize results
steroid_summary <- lapply(seq_along(steroid_model), function(i) {
  model <- steroid_model[[i]]
  feature <- all_features[i]
  coef_summary <- summary(model)$coefficients
  steroid_row <- rownames(coef_summary) == "steroids_days"

  data.frame(
    feature = feature,
    OR = exp(coef_summary[steroid_row, "Estimate"]),
    lowerCI = exp(coef_summary[steroid_row, "Estimate"] - 1.96 * coef_summary[steroid_row, "Std. Error"]),
    upperCI = exp(coef_summary[steroid_row, "Estimate"] + 1.96 * coef_summary[steroid_row, "Std. Error"]),
    pvalue = coef_summary[steroid_row, "Pr(>|z|)"],
    group0 = sum(combined[[feature]] == 0, na.rm = TRUE),
    group1 = sum(combined[[feature]] == 1, na.rm = TRUE)
  )
})

# Combine and adjust p-values
steroid_summary <- do.call(rbind, steroid_summary)
steroid_summary$AdjP <- p.adjust(steroid_summary$pvalue, method = "fdr")
```

<details><summary><strong>View steriod expsure model results</strong></summary>
```{r, echo=FALSE}
knitr::kable(steroid_summary, caption = "Steriod Exposure Model Results")
```
</details>

```{r, echo=FALSE, fig.width=12, fig.height=7}
# Prepare base data frame with odds ratios
base_data <- tibble(
  mean  = steroid_summary$OR,       # OR
  lower = steroid_summary$lowerCI,        # Lower 95% CI
  upper = steroid_summary$upperCI,        # Upper 95% CI
  feature = steroid_summary$feature,
  pvalue_raw = steroid_summary$pvalue,
  pvalue_adj = steroid_summary$AdjP,
  estimate_ci = paste0(
    sprintf("%.2f", steroid_summary$OR), " [",
    sprintf("%.2f", steroid_summary$lowerCI), ", ",
    sprintf("%.2f", steroid_summary$upperCI), "]"
  ),
  group0 = steroid_summary$group0,
  group1 = steroid_summary$group1

) %>%
  mutate(
    feature_clean = str_replace_all(feature, "_", " "),
    feature_clean = str_remove(feature_clean, " ?bin$"),
    feature_clean = str_to_title(feature_clean),
    
    pvalue_str = sprintf("%.3f", pvalue_raw),
    adj_pvalue_str = sprintf("%.3f", pvalue_adj),
    sig_star = ifelse(pvalue_raw < 0.05, "*", ""),
    feature_label = paste0(feature_clean, sig_star),
    sample_sizes = sapply(steroid_summary$feature, function(f) {sum(!is.na(combined[[f]]))}),
    group0 = paste(group0,"/", sample_sizes, sep = ""),
    group1 = paste(group1,"/", sample_sizes, sep = "")

  )

# Create label text
labeltext <- rbind(
  c("Feature", "Negative", "Positive", "OR [95% CI]", "P-Value", "Adj. P-Value"),
  as.matrix(cbind(
    base_data$feature_label,
    base_data$group0,
    base_data$group1,
    base_data$estimate_ci,
    base_data$pvalue_str,
    base_data$adj_pvalue_str  ))
)

# Add NA to align rows with label header
mean_all  <- c(NA, base_data$mean)
lower_all <- c(NA, base_data$lower)
upper_all <- c(NA, base_data$upper)


# Forest plot
forestplot(
  labeltext = labeltext,
  mean = mean_all,
  lower = lower_all,
  upper = upper_all,
  is.summary = c(TRUE, rep(FALSE, nrow(base_data))),
  boxsize = 0.2,
  col = fpColors(box = "slateblue4", line = "mediumorchid4", zero = "red"),
  xlog = TRUE,  # log scale for odds ratios
  clip = c(min(base_data$lower, na.rm = TRUE), max(base_data$upper, na.rm = TRUE)),
  xticks = pretty(c(base_data$lower, base_data$upper)),
  title = "Impact of Steroid Exposure on Vascular Histopathology",
  txt_gp = fpTxtGp(
    label = gpar(cex = 0.8),
    ticks = gpar(cex = 0.8),
    xlab = gpar(cex = 0.9)
  ))

grid.text("*P-values adjusted with Benjamini-Hochberg (FDR) procedure", x = 0.01, y = unit(0.1, "npc") - unit(2, "lines"), 
          just = "left", gp = gpar(cex = 0.8, col = "black"))

```


#### 3.2.2 Sex Models
___
We assessed whether patient sex (male vs. female) was associated with histological features using logistic regression.

```{r, echo=FALSE}
combined$sex_binary <- ifelse(combined$sex == 2, 1, 
                              ifelse(combined$sex == 1, 0, NA))

# Fit logistic models: feature ~ sex_binary
sex_model <- lapply(all_features, function(feature) {
  glm(as.formula(paste(feature, "~ sex_binary")), data = combined, family = binomial)
})

# Name the models by feature
names(sex_model) <- all_features

# Extract effect estimates, CIs, and p-values
sex_summary <- lapply(seq_along(sex_model), function(i) {
  model <- sex_model[[i]]
  feature <- all_features[i]

  # Get coefficient summary safely
  coef_summary <- tryCatch(summary(model)$coefficients, error = function(e) NULL)

  # Skip if model failed or sex_binary is not in the result
  if (is.null(coef_summary) || !"sex_binary" %in% rownames(coef_summary)) {
    return(NULL)
  }

  data.frame(
    feature = feature,
    OR = exp(coef_summary["sex_binary", "Estimate"]),
    lowerCI = exp(coef_summary["sex_binary", "Estimate"] - 1.96 * coef_summary["sex_binary", "Std. Error"]),
    upperCI = exp(coef_summary["sex_binary", "Estimate"] + 1.96 * coef_summary["sex_binary", "Std. Error"]),
    pvalue = coef_summary["sex_binary", "Pr(>|z|)"],
    group0 = sum(combined[[feature]] == 0, na.rm = TRUE),
    group1 = sum(combined[[feature]] == 1, na.rm = TRUE)
  )
})

# Remove any NULLs (failed models)
sex_summary <- do.call(rbind, sex_summary)
sex_summary$AdjP <- p.adjust(sex_summary$pvalue, method = "fdr")
```

<details><summary><strong>View sex model results</strong></summary>
```{r, echo=FALSE}
knitr::kable(sex_summary, caption = "Sex Model Results")
```
</details>

```{r, echo=FALSE, fig.width=12, fig.height=7}
sex_base_data <- tibble(
  mean  = sex_summary$OR,
  lower = sex_summary$lowerCI,
  upper = sex_summary$upperCI,
  feature = sex_summary$feature,
  pvalue_raw = sex_summary$pvalue,
  pvalue_adj = sex_summary$AdjP,
  estimate_ci = paste0(
    sprintf("%.2f", sex_summary$OR), " [",
    sprintf("%.2f", sex_summary$lowerCI), ", ",
    sprintf("%.2f", sex_summary$upperCI), "]"
  ),
  group0 = steroid_summary$group0,
  group1 = steroid_summary$group1
) %>%
  mutate(
    feature_clean = str_replace_all(feature, "_", " "),
    feature_clean = str_remove(feature_clean, " ?bin$"),
    feature_clean = str_to_title(feature_clean),
    pvalue_str = sprintf("%.3f", pvalue_raw),
    adj_pvalue_str = sprintf("%.3f", pvalue_adj),
    sig_star = ifelse(pvalue_raw < 0.05, "*", ""),
    feature_label = paste0(feature_clean, sig_star),
    sample_sizes = sapply(sex_summary$feature, function(f) {sum(!is.na(combined[[f]]))}),
    group0 = paste(group0,"/", sample_sizes, sep = ""),
    group1 = paste(group1,"/", sample_sizes, sep = "")

  )


# Filter NA or problematic values
sex_base_data <- sex_base_data %>%
  filter(!is.na(mean) & !is.na(lower) & !is.na(upper))

# Label matrix
sex_labeltext <- rbind(
  c("Feature", "Negative", "Positive", "OR [95% CI]", "P-Value", "Adj. P-Value"),
  as.matrix(cbind(
    sex_base_data$feature_label,
    base_data$group0,
    base_data$group1,
    sex_base_data$estimate_ci,
    sex_base_data$pvalue_str,
    sex_base_data$adj_pvalue_str
  ))
)

# Prepare values
mean_all  <- c(NA, sex_base_data$mean)
lower_all <- c(NA, sex_base_data$lower)
upper_all <- c(NA, sex_base_data$upper)


# Plot on linear scale
forestplot(
  labeltext = sex_labeltext,
  mean = mean_all,
  lower = lower_all,
  upper = upper_all,
  is.summary = c(TRUE, rep(FALSE, nrow(sex_base_data))),
  boxsize = 0.2,
  col = fpColors(box =  "royalblue", line = "darkblue", zero = "red"),
  xlog = FALSE,  # linear scale for log-odds
  clip = c(min(lower_all, na.rm = TRUE), max(upper_all, na.rm = TRUE)),
  xticks = pretty(c(min(lower_all, na.rm = TRUE), max(upper_all, na.rm = TRUE))),
  title = "Influence of Patient Sex on Histological Features",
  txt_gp = fpTxtGp(
    label = gpar(cex = 0.8),
    ticks = gpar(cex = 0.8),
    xlab = gpar(cex = 0.9)
  )
)

grid.text("*P-values adjusted with Benjamini-Hochberg (FDR) procedure", x = 0.01, y = unit(0.1, "npc") - unit(2, "lines"), 
          just = "left", gp = gpar(cex = 0.8, col = "black"))



```


#### 3.2.3 Age Models
___
Next, we modeled each binary histological feature as a function of age.

```{r, echo=FALSE}
# Loop over binary outcome features
age_model <- lapply(all_features, function(feature) {
  glm(as.formula(paste(feature, "~ age")), data = combined, family = binomial)
})

names(age_model) <- all_features

# Extract ORs, CIs, and p-values
age_summary <- lapply(seq_along(age_model), function(i) {
  model <- age_model[[i]]
  feature <- all_features[i]

  coef_summary <- tryCatch(summary(model)$coefficients, error = function(e) NULL)

  if (is.null(coef_summary) || !"age" %in% rownames(coef_summary)) {
    return(NULL)
  }

  est <- coef_summary["age", "Estimate"]
  se <- coef_summary["age", "Std. Error"]
  pval <- coef_summary["age", "Pr(>|z|)"]

  data.frame(
    Feature = feature,
    Predictor = "Age",
    OR = exp(est),
    LowerCI = exp(est - 1.96 * se),
    UpperCI = exp(est + 1.96 * se),
    PValue = pval,
    group0 = sum(combined[[feature]] == 0, na.rm = TRUE),
    group1 = sum(combined[[feature]] == 1, na.rm = TRUE)
  )
})

# Combine and adjust p-values
age_summary <- do.call(rbind, age_summary)
age_summary$AdjP <- p.adjust(age_summary$PValue, method = "fdr")

```

<details><summary><strong>View age model results</strong></summary>
```{r, echo=FALSE}
knitr::kable(age_summary, caption = "Age Model Results")
```
</details>

```{r, echo=FALSE, fig.width=12, fig.height=7}
# Prepare the data for forestplot
age_base_data <- tibble(
  mean  = age_summary$OR,
  lower = age_summary$LowerCI,
  upper = age_summary$UpperCI,
  feature = age_summary$Feature,
  pvalue_raw = age_summary$PValue,
  pvalue_adj = age_summary$AdjP,
  estimate_ci = paste0(
    sprintf("%.2f", age_summary$OR), " [",
    sprintf("%.2f", age_summary$LowerCI), ", ",
    sprintf("%.2f", age_summary$UpperCI), "]"
  ),
  group0 = steroid_summary$group0,
  group1 = steroid_summary$group1
) %>%
  mutate(
    # Clean and label features
    feature_clean = str_replace_all(feature, "_", " "),
    feature_clean = str_remove(feature_clean, " ?bin$"),
    feature_clean = str_to_title(feature_clean),

    # Format p-values
    pvalue_str = sprintf("%.3f", pvalue_raw),
    adj_pvalue_str = sprintf("%.3f", pvalue_adj),
    sig_star = ifelse(pvalue_raw < 0.05, "*", ""),
    feature_label = paste0(feature_clean, sig_star),
    sample_sizes = sapply(age_summary$Feature, function(f) {sum(!is.na(combined[[f]]))}),
    group0 = paste(group0,"/", sample_sizes, sep = ""),
    group1 = paste(group1,"/", sample_sizes, sep = "")
  )


# Prepare the label text for the forest plot
age_labeltext <- rbind(
  c("Feature", "Negative", "Positive", "OR [95% CI]", "P-Value", "Adj. P-Value"),
  as.matrix(cbind(
    age_base_data$feature_label,
    base_data$group0,
    base_data$group1,
    age_base_data$estimate_ci,
    age_base_data$pvalue_str,
    age_base_data$adj_pvalue_str
    ))
)


# Add NA to align with the header row
mean_all  <- c(NA, age_base_data$mean)
lower_all <- c(NA, age_base_data$lower)
upper_all <- c(NA, age_base_data$upper)

# Safe clipping range
min_val <- max(min(age_base_data$lower, na.rm = TRUE), 0.01)
max_val <- max(age_base_data$upper, na.rm = TRUE)


# Forest plot
forestplot(
  labeltext = age_labeltext,
  mean = mean_all,
  lower = lower_all,
  upper = upper_all,
  is.summary = c(TRUE, rep(FALSE, nrow(age_base_data))),
  boxsize = 0.2,
  col = fpColors(box = "darkgreen", line = "forestgreen", zero = "red"),
  xlog = TRUE,
  clip = c(min_val, max_val),
  xticks = pretty(c(min_val, max_val)),
  title = "Impact of Age on Vascular Histopathology",
  txt_gp = fpTxtGp(
    label = gpar(cex = 0.8),
    ticks = gpar(cex = 0.8),
    xlab  = gpar(cex = 0.9)
  )
)

grid.text("*P-values adjusted with Benjamini-Hochberg (FDR) procedure", x = 0.01, y = unit(0.1, "npc") - unit(2, "lines"), 
          just = "left", gp = gpar(cex = 0.8, col = "black"))

```


### 3.3 Interpretation of Results
Across all three models (age, sex, and steroid exposure), no histological features reached statistical significance. These results indicate that, within this cohort, these demographic and treatment variables are not associated with variation in histopathological features and may not serve as meaningful covariates in subsequent analyses.

## 4. Differential Gene Expression

Comparisons Include: True negative controls (GCA = 0) vs. steroid exposure of

- less than 3 days (0, 1, 2 days)
- 3-4 days
- 5-7 days
- more than 7 days

```{r, echo=FALSE}
gene_data <-read.table("/Users/kristinemedcalf/Desktop/Programming/Dissertation_GCA_Prediction/data/matrix_salmon_tximport_counts_cc_c1_c2_GENCODE.txt")

gene_data <- round(as.matrix(gene_data))  # round to nearest integer
mode(gene_data) <- "numeric"              # ensure numeric mode
# need to ask if raw counts? cause why decimal?
clin_var <- read.csv('/Users/kristinemedcalf/Desktop/Programming/Dissertation_GCA_Prediction/data/clinical_variables_cc_c1_c2.csv')
hist_var <- read.csv("/Users/kristinemedcalf/Desktop/Programming/Dissertation_GCA_Prediction/data/histological_variables_cc_c1_c2.csv")

combined <- merge(clin_var, hist_var)

combined <- combined %>%
  mutate(steroid_group = case_when(
    is.na(steroids_days)         ~ NA_character_,
    GCA_present == 0             ~ "true_negative",
    steroids_days < 3            ~ "less_3",
    steroids_days < 5            ~ "days_3_4",
    steroids_days < 8            ~ "days_5_6_7",
    steroids_days >= 8           ~ "more_7"
  ))

```


### 4.1 Steriod Exposure of Less than 3 days

```{r, echo=FALSE, message=FALSE}

df <- combined %>%
  filter(steroid_group %in% c("less_3", "true_negative"))

col_ids <- str_sub(colnames(gene_data), 2)
df_filtered <- df[df$ID %in% col_ids, ]

df_filtered <- df_filtered[!is.na(df_filtered$steroid_group), ]
gene_data_filtered <- gene_data[, paste0("X", df_filtered$ID)]

coldata <- data.frame(
  row.names = paste0("X", df_filtered$ID),
  steroid_group = factor(df_filtered$steroid_group, levels = c("true_negative", "less_3"))
)

dds <- DESeqDataSetFromMatrix(
  countData = gene_data_filtered,
  colData = coldata,
  design = ~ steroid_group
)
dds <- DESeq(dds)

res <- results(dds, contrast = c("steroid_group", "less_3", "true_negative"))

res_df <- as.data.frame(res)
res_df$gene <- rownames(res_df)

res_df$ensembl_clean <- gsub("\\..*", "", res_df$gene)

ensembl <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")
annotations <- getBM(
  attributes = c("ensembl_gene_id", "hgnc_symbol", "description"),
  filters = "ensembl_gene_id",
  values = res_df$ensembl_clean,
  mart = ensembl
)

res_df_annotated <- merge(
  res_df,
  annotations,
  by.x = "ensembl_clean",
  by.y = "ensembl_gene_id",
  all.x = TRUE
)

sig_genes <- res_df_annotated[!is.na(res_df_annotated$padj) & res_df_annotated$padj < 0.05, ]
sig_genes <- sig_genes[order(sig_genes$padj), ]

foldchange_genes <- sig_genes[rev(order(abs(sig_genes$log2FoldChange))), ]
```

<details><summary><strong>View Results Ordered By Significance </strong></summary>
```{r, echo=FALSE}
knitr::kable(sig_genes[, c('gene', 'hgnc_symbol', 'log2FoldChange', 'padj' )], caption = "Steroid Exposure Less than 3 Day ")
```
</details>

<details><summary><strong>View Results Ordered By Log2FoldChange </strong></summary>
```{r, echo=FALSE}
knitr::kable(foldchange_genes[, c('gene', 'hgnc_symbol', 'log2FoldChange', 'padj' )], caption = "Steroid Exposure Less than 3 Day ")
```
</details>

### 4.2 Steriod Exposure Between 3 to 4 Days

```{r, echo=FALSE, message=FALSE}

df <- combined %>%
  filter(steroid_group %in% c("days_3_4", "true_negative"))

col_ids <- str_sub(colnames(gene_data), 2)
df_filtered <- df[df$ID %in% col_ids, ]

df_filtered <- df_filtered[!is.na(df_filtered$steroid_group), ]

gene_data_filtered <- gene_data[, paste0("X", df_filtered$ID)]

coldata <- data.frame(
  row.names = paste0("X", df_filtered$ID),
  steroid_group = factor(df_filtered$steroid_group, levels = c("true_negative", "days_3_4"))
)

dds <- DESeqDataSetFromMatrix(
  countData = gene_data_filtered,
  colData = coldata,
  design = ~ steroid_group
)
dds <- DESeq(dds)

res <- results(dds, contrast = c("steroid_group", "days_3_4", "true_negative"))

res_df <- as.data.frame(res)
res_df$gene <- rownames(res_df)

res_df$ensembl_clean <- gsub("\\..*", "", res_df$gene)

ensembl <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")
annotations <- getBM(
  attributes = c("ensembl_gene_id", "hgnc_symbol", "description"),
  filters = "ensembl_gene_id",
  values = res_df$ensembl_clean,
  mart = ensembl
)

res_df_annotated <- merge(
  res_df,
  annotations,
  by.x = "ensembl_clean",
  by.y = "ensembl_gene_id",
  all.x = TRUE
)

sig_genes <- res_df_annotated[!is.na(res_df_annotated$padj) & res_df_annotated$padj < 0.05, ]
sig_genes <- sig_genes[order(sig_genes$padj), ]


foldchange_genes <- sig_genes[rev(order(abs(sig_genes$log2FoldChange))), ]
```

<details><summary><strong>View Results Ordered By Significance </strong></summary>
```{r, echo=FALSE}
knitr::kable(sig_genes[, c('gene', 'hgnc_symbol', 'log2FoldChange', 'padj' )], caption = "Steroid Exposure Less than 3 Day ")
```
</details>

<details><summary><strong>View Results Ordered By Log2FoldChange </strong></summary>
```{r, echo=FALSE}
knitr::kable(foldchange_genes[, c('gene', 'hgnc_symbol', 'log2FoldChange', 'padj' )], caption = "Steroid Exposure Less than 3 Day ")
```
</details>

### 4.3 Steriod Exposure Between 5 to 7 Days

```{r, echo=FALSE, message=FALSE}

df <- combined %>%
  filter(steroid_group %in% c("days_5_6_7", "true_negative"))

col_ids <- str_sub(colnames(gene_data), 2)
df_filtered <- df[df$ID %in% col_ids, ]

df_filtered <- df_filtered[!is.na(df_filtered$steroid_group), ]

gene_data_filtered <- gene_data[, paste0("X", df_filtered$ID)]

coldata <- data.frame(
  row.names = paste0("X", df_filtered$ID),
  steroid_group = factor(df_filtered$steroid_group, levels = c("true_negative", "days_5_6_7"))
)

dds <- DESeqDataSetFromMatrix(
  countData = gene_data_filtered,
  colData = coldata,
  design = ~ steroid_group
)
dds <- DESeq(dds)

res <- results(dds, contrast = c("steroid_group", "days_5_6_7", "true_negative"))

res_df <- as.data.frame(res)
res_df$gene <- rownames(res_df)

res_df$ensembl_clean <- gsub("\\..*", "", res_df$gene)

ensembl <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")
annotations <- getBM(
  attributes = c("ensembl_gene_id", "hgnc_symbol", "description"),
  filters = "ensembl_gene_id",
  values = res_df$ensembl_clean,
  mart = ensembl
)

res_df_annotated <- merge(
  res_df,
  annotations,
  by.x = "ensembl_clean",
  by.y = "ensembl_gene_id",
  all.x = TRUE
)

sig_genes <- res_df_annotated[!is.na(res_df_annotated$padj) & res_df_annotated$padj < 0.05, ]
sig_genes <- sig_genes[order(sig_genes$padj), ]

foldchange_genes <- sig_genes[rev(order(abs(sig_genes$log2FoldChange))), ]
```

<details><summary><strong>View Results Ordered By Significance </strong></summary>
```{r, echo=FALSE}
knitr::kable(sig_genes[, c('gene', 'hgnc_symbol', 'log2FoldChange', 'padj' )], caption = "Steroid Exposure Less than 3 Day ")
```
</details>

<details><summary><strong>View Results Ordered By Log2FoldChange </strong></summary>
```{r, echo=FALSE}
knitr::kable(foldchange_genes[, c('gene', 'hgnc_symbol', 'log2FoldChange', 'padj' )], caption = "Steroid Exposure Less than 3 Day ")
```
</details>

### 4.4 Steriod Exposure of More than 7 Days

```{r, echo=FALSE, message=FALSE}

df <- combined %>%
  filter(steroid_group %in% c("more_7", "true_negative"))

col_ids <- str_sub(colnames(gene_data), 2)
df_filtered <- df[df$ID %in% col_ids, ]

df_filtered <- df_filtered[!is.na(df_filtered$steroid_group), ]

gene_data_filtered <- gene_data[, paste0("X", df_filtered$ID)]

coldata <- data.frame(
  row.names = paste0("X", df_filtered$ID),
  steroid_group = factor(df_filtered$steroid_group, levels = c("true_negative", "more_7"))
)

dds <- DESeqDataSetFromMatrix(
  countData = gene_data_filtered,
  colData = coldata,
  design = ~ steroid_group
)
dds <- DESeq(dds)

res <- results(dds, contrast = c("steroid_group", "more_7", "true_negative"))

res_df <- as.data.frame(res)
res_df$gene <- rownames(res_df)

res_df$ensembl_clean <- gsub("\\..*", "", res_df$gene)

ensembl <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")
annotations <- getBM(
  attributes = c("ensembl_gene_id", "hgnc_symbol", "description"),
  filters = "ensembl_gene_id",
  values = res_df$ensembl_clean,
  mart = ensembl
)

res_df_annotated <- merge(
  res_df,
  annotations,
  by.x = "ensembl_clean",
  by.y = "ensembl_gene_id",
  all.x = TRUE
)

sig_genes <- res_df_annotated[!is.na(res_df_annotated$padj) & res_df_annotated$padj < 0.05, ]
sig_genes <- sig_genes[order(sig_genes$padj), ]

foldchange_genes <- sig_genes[rev(order(abs(sig_genes$log2FoldChange))), ]
```

<details><summary><strong>View Results Ordered By Significance </strong></summary>
```{r, echo=FALSE}
knitr::kable(sig_genes[, c('gene', 'hgnc_symbol', 'log2FoldChange', 'padj' )], caption = "Steroid Exposure Less than 3 Day ")
```
</details>

<details><summary><strong>View Results Ordered By Log2FoldChange </strong></summary>
```{r, echo=FALSE}
knitr::kable(foldchange_genes[, c('gene', 'hgnc_symbol', 'log2FoldChange', 'padj' )], caption = "Steroid Exposure Less than 3 Day ")
```
</details>

```{r}
library(pheatmap)
ntd <- vst(dds, blind = TRUE)
top_genes <- head(foldchange_genes$gene, 20)
mat <- assay(ntd)[top_genes, ]
gene_symbols <- foldchange_genes$hgnc_symbol[match(top_genes, foldchange_genes$gene)]
rownames(mat) <- ifelse(is.na(gene_symbols), top_genes, gene_symbols)
pheatmap(mat,
         annotation_col = coldata,
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         show_rownames = TRUE,
         scale = "row")
```


