---
title: "Impact_GCA"
header-includes:
   - \usepackage{amsmath}
   - \usepackage{subcaption}
   - \usepackage{fullpage}
   - \usepackage{textcomp}
   - \usepackage{graphicx}
   - \usepackage{sidecap}
   - \usepackage{setspace}
   - \usepackage{sectsty}
   - \usepackage{listings}
   - \usepackage{amssymb}
   - \newcommand{\iput}{\textrm{\textsc{input}}}
   - \newcommand{\oput}{\textrm{\textsc{output}}}
   - \newcommand{\N}{\mathcal{N}}
   - \newcommand{\Lik}{\mathcal{L}}
   - \newcommand{\q}{\textbf{Q.\stepcounter{t}\thet}}
output:
  pdf_document:
    number_sections: true
editor_options: 
  markdown: 
    wrap: sentence
---

# First Objective: Assess Effect of Steroids on GCA
One is with clinical variables and the other with histological phenotypes. To address the first objective of your project, which was to assess the effect of steroids and other clinical variables on the phenotypes of interest,  you can run regression analysis using the histological phenotypes as outcome and clinical variables as exposure.

I’d suggest to just pick one of the histological phenotypes, e.g. “Media_destruction” and run three separate regression models for these three exposures: duration of steroids, sex and age. Please let me know when you get some preliminary results, and we can then discuss further, and you can run the analysis analogously for other histological and clinical phenotypes of interest.

Aim 1: Effect of clinical variables on histological phenotypes
Aim 2: Effect of clinical variables on diagnosis (e.g. GCA_present)

## Import Files
```{r}
library(ggplot2)
clinical <- read.csv('/Users/kristinemedcalf/Desktop/Programming/Dissertation_GCA_Prediction/data/clinical_variables_cc_c1_c2.csv')
hist_var <- read.csv("/Users/kristinemedcalf/Desktop/Programming/Dissertation_GCA_Prediction/data/histological_variables_cc_c1_c2.csv")

# merge together
combined <- merge(clinical, hist_var, by = "ID")
phenotypes <- names(hist_var)[-1]
clin_var <- names(clinical)[-1]
num_phenotypes <- length(phenotypes)
```

## Heatmap
```{r}
cor_matrix <- cor(combined[, phenotypes], use = "pairwise.complete.obs")
library(pheatmap)
pheatmap(cor_matrix, cluster_rows = TRUE, cluster_cols = TRUE)

numeric_clin <- combined[, clin_var][, sapply(combined[, clin_var], is.numeric)]
numeric_clin <- numeric_clin[, apply(numeric_clin, 2, function(x) var(x, na.rm = TRUE) != 0)]
cor_matrix <- cor(numeric_clin, use = "pairwise.complete.obs")
library(pheatmap)
pheatmap(cor_matrix, display_numbers = TRUE, cluster_rows = TRUE, cluster_cols = TRUE)

# dont use spearman/pearson
```

```{r}
cor_results <- data.frame(
  phenotype = character(),
  correlation = numeric(),
  p_value = numeric(),
  stringsAsFactors = FALSE
)

for (pheno in phenotypes) {
  x <- combined[[pheno]]
  y <- combined$GCA_present
  if (is.numeric(x) && length(na.omit(x)) > 2) {
    test <- cor.test(x, y, method = "pearson", use = "complete.obs")
    cor_results <- rbind(cor_results, data.frame(
      phenotype = pheno,
      correlation = test$estimate,
      p_value = test$p.value
    ))
  }
}

cor_results <- cor_results[order(-abs(cor_results$correlation)), ]
print(cor_results)

```

```{r}
library(ggplot2)

ggplot(cor_results, aes(x = reorder(phenotype, correlation), y = correlation)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  geom_hline(yintercept = 0, color = "black") +
  coord_flip() +
  theme_minimal() +
  labs(title = "Correlation Between Histological Phenotypes and GCA Diagnosis",
       x = "Phenotype", y = "Pearson Correlation Coefficient")

```

Blocks of high correlation --> suggest many histological features co-occur.

GCA_present is moderately to strongly correlated with many features. This suggests many of these phenotypes track with GCA diagnosis -> they are histological hallmarks or correlates of GCA.

Good for multicollinearity issues. 

## Simple Regression Models

```{r}
steroid_models <- vector("list", num_phenotypes)
age_models <- vector("list", num_phenotypes)
sex_models <- vector("list", num_phenotypes)
```

### Linear Regression
```{r}
count <- 0
for (variable in phenotypes) {
  count <- count + 1
  
  # sex
  formula <- as.formula(paste(variable, "~ sex"))
  model_sex <- lm(formula, data = combined)
  sex_models[[count]] <-model_sex 
 
  # steroids
  formula <- as.formula(paste(variable, "~ steroids_days"))
  model_steroids <- lm(formula, data = combined)
  steroid_models[[count]] <-model_steroids 
  
  # age
  formula <- as.formula(paste(variable, "~ age"))
  model_age <- lm(formula, data = combined)
  age_models[[count]] <-model_age 
}

names(sex_models) <- phenotypes
names(age_models) <- phenotypes
names(steroid_models) <- phenotypes
```

Get p-values
```{r}
pvals_sex <- numeric(length(phenotypes))
pvals_age <- numeric(length(phenotypes))
pvals_steroids <- numeric(length(phenotypes))

for (i in seq_along(phenotypes)) {
  # For sex
  pvals_sex[i] <- summary(sex_models[[i]])$coefficients["sex", "Pr(>|t|)"]
  
  # For age
  pvals_age[i] <- summary(age_models[[i]])$coefficients["age", "Pr(>|t|)"]
  
  # For steroids
  pvals_steroids[i] <- summary(steroid_models[[i]])$coefficients["steroids_days", "Pr(>|t|)"]
}

# pvals_sex <- p.adjust(pvals_sex, method = "fdr")
# pvals_age <- p.adjust(pvals_age, method = "fdr")
# pvals_steroids <- p.adjust(pvals_steroids, method = "fdr")

# plot
df <- data.frame(
  phenotype = phenotypes,
  pval = pvals_sex
)

ggplot(df, aes(x = phenotype, y = -log10(pval))) +
  geom_bar(stat = "identity", fill = "skyblue") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "red") +
  theme_get() +
  theme(
    axis.text.x = element_text(angle = 60, hjust = 1, size = 8),
    plot.margin = margin(10, 10, 40, 10)
  ) +
  labs(title = "Significant Phenotypes Based on Sex", x = "Phenotype", y = "-log10(p)")


df <- data.frame(
  phenotype = phenotypes,
  pval = pvals_age
)

ggplot(df, aes(x = phenotype, y = -log10(pval))) +
  geom_bar(stat = "identity", fill = "skyblue") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "red") +
  theme_get() +
  theme(
    axis.text.x = element_text(angle = 60, hjust = 1, size = 8),
    plot.margin = margin(10, 10, 40, 10)
  ) +
  labs(title = "Significant Phenotypes Based on Age", x = "Phenotype", y = "-log10(p)")


df <- data.frame(
  phenotype = phenotypes,
  pval = pvals_steroids
)

ggplot(df, aes(x = phenotype, y = -log10(pval))) +
  geom_bar(stat = "identity", fill = "skyblue") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "red") +
  theme_get() +
  theme(
    axis.text.x = element_text(angle = 60, hjust = 1, size = 8),
    plot.margin = margin(10, 10, 40, 10)
  ) +
  labs(title = "Significant Phenotypes Based on Steriod Exposure (in Days)", x = "Phenotype", y = "-log10(p)")

```
break age into 5 year blocks
Effect size (beta coeffecients) and confidence intervals across all histological phenotypes

```{r}
# Initialize storage
results_df <- data.frame(
  phenotype = phenotypes,
  beta_age = NA, ci_lower_age = NA, ci_upper_age = NA,
  beta_sex = NA, ci_lower_sex = NA, ci_upper_sex = NA,
  beta_steroids = NA, ci_lower_steroids = NA, ci_upper_steroids = NA
)

# Loop through each phenotype
for (i in seq_along(phenotypes)) {
  var <- phenotypes[i]
  
  # Age model
  model_age <- age_models[[i]]
  if (!is.null(model_age)) {
    coef_age <- summary(model_age)$coefficients
    ci_age <- confint(model_age)
    results_df$beta_age[i] <- coef_age["age", "Estimate"]
    results_df$ci_lower_age[i] <- ci_age["age", 1]
    results_df$ci_upper_age[i] <- ci_age["age", 2]
  }
  
  # Sex model
  model_sex <- sex_models[[i]]
  if (!is.null(model_sex)) {
    coef_sex <- summary(model_sex)$coefficients
    ci_sex <- confint(model_sex)
    results_df$beta_sex[i] <- coef_sex["sex", "Estimate"]
    results_df$ci_lower_sex[i] <- ci_sex["sex", 1]
    results_df$ci_upper_sex[i] <- ci_sex["sex", 2]
  }
  
  # Steroids model
  model_steroids <- steroid_models[[i]]
  if (!is.null(model_steroids)) {
    coef_steroids <- summary(model_steroids)$coefficients
    ci_steroids <- confint(model_steroids)
    results_df$beta_steroids[i] <- coef_steroids["steroids_days", "Estimate"]
    results_df$ci_lower_steroids[i] <- ci_steroids["steroids_days", 1]
    results_df$ci_upper_steroids[i] <- ci_steroids["steroids_days", 2]
  }
}
# Plot for Age
ggplot(results_df, aes(x = reorder(phenotype, beta_age), y = beta_age)) +
  geom_point(color = "steelblue") +
  geom_errorbar(aes(ymin = ci_lower_age, ymax = ci_upper_age), width = 0.2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  theme_minimal() +
  coord_flip() +
  labs(title = "Effect of Age on Histological Phenotypes",
       x = "Phenotype",
       y = "Beta Coefficient (with 95% CI)")

# Plot for Sex
ggplot(results_df, aes(x = reorder(phenotype, beta_sex), y = beta_sex)) +
  geom_point(color = "steelblue") +
  geom_errorbar(aes(ymin = ci_lower_sex, ymax = ci_upper_sex), width = 0.2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  theme_minimal() +
  coord_flip() +
  labs(title = "Effect of Sex on Histological Phenotypes",
       x = "Phenotype",
       y = "Beta Coefficient (with 95% CI)")

# Plot for Steroid Exposure
ggplot(results_df, aes(x = reorder(phenotype, beta_steroids), y = beta_steroids)) +
  geom_point(color = "steelblue") +
  geom_errorbar(aes(ymin = ci_lower_steroids, ymax = ci_upper_steroids), width = 0.2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  theme_minimal() +
  coord_flip() +
  labs(title = "Effect of Steroid Exposure on Histological Phenotypes",
       x = "Phenotype",
       y = "Beta Coefficient (with 95% CI)")

```

### Logistic Regression Model
Instead of linearly --> look at steroids via logistic regression
```{r}
library(logistf)

phenotypes_log <- setdiff(names(combined), "GCA_present")

firth_models <- vector("list", length(phenotypes_log))
pvals_clinical <- numeric(length(phenotypes_log))

for (i in seq_along(phenotypes_log)) {
  variable <- phenotypes_log[i]
  formula <- as.formula(paste("GCA_present ~", variable))
  
  # Clean data for this phenotype
  subset_data <- combined[, c("GCA_present", variable)]
  subset_data <- subset_data[complete.cases(subset_data), ]
  
  if (nrow(subset_data) < 10 || length(unique(subset_data$GCA_present)) < 2) {
    pvals_clinical[i] <- NA
    next
  }
  
  model <- tryCatch(
    logistf(formula, data = subset_data),
    error = function(e) NULL
  )
  
  firth_models[[i]] <- model
  
  # Extract p-value if model ran
  if (!is.null(model) && length(model$prob) > 0) {
    pvals_clinical[i] <- model$prob[2]  # second row = your phenotype
  } else {
    pvals_clinical[i] <- NA
  }
}

names(pvals_clinical) <- phenotypes_log
pvals_fdr <- p.adjust(pvals_clinical, method = "fdr")

```

```{r}
df <- data.frame(
  phenotype = phenotypes_log,
  pval = pvals_fdr
)
df <- df[!is.na(df$pval), ]

ggplot(df, aes(x = reorder(phenotype, -log10(pval)), y = -log10(pval))) +
  geom_bar(stat = "identity", fill = "skyblue") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "red") +
  coord_flip() +
  theme_minimal() +
  labs(title = "Firth Logistic Regression: Phenotypes Predicting GCA",
       x = "Histological Phenotype", y = "-log10(p-value)")

```
Logistic regression was difficult because many variables perfectly seperated. Used Firth's regression in order to penalise likelihood to allow regression. 


## Multivariable Regression Models

### Linear Regression Model
```{r}
# Initialize storage
multi_results <- data.frame(
  phenotype = character(),
  variable = character(),
  beta = numeric(),
  ci_lower = numeric(),
  ci_upper = numeric(),
  p_value = numeric(),
  stringsAsFactors = FALSE
)

# Loop through each histological phenotype
for (pheno in phenotypes) {
  formula <- as.formula(paste(pheno, "~ sex + age + steroids_days"))
  model <- lm(formula, data = combined)
  
  # Get summary and confidence intervals
  coefs <- summary(model)$coefficients
  conf <- confint(model)
  
  for (var in c("sex", "age", "steroids_days")) {
    if (var %in% rownames(coefs)) {
      multi_results <- rbind(multi_results, data.frame(
        phenotype = pheno,
        variable = var,
        beta = coefs[var, "Estimate"],
        ci_lower = conf[var, 1],
        ci_upper = conf[var, 2],
        p_value = coefs[var, "Pr(>|t|)"]
      ))
    }
  }
}

```

```{r}
significant_results <- subset(multi_results, p_value < 0.05)

plot_effects <- function(df, predictor) {
  df_subset <- subset(df, variable == predictor)
  ggplot(df_subset, aes(x = reorder(phenotype, beta), y = beta)) +
    geom_point(color = "steelblue") +
    geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.2) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
    coord_flip() +
    theme_minimal() +
    labs(
      title = paste("Adjusted Effect of", predictor, "on Histological Phenotypes"),
      x = "Phenotype",
      y = "Beta Coefficient (95% CI)"
    )
}

plot_effects(multi_results, "age")
plot_effects(multi_results, "sex")
plot_effects(multi_results, "steroids_days")

```
keep adventitia, drop any
```{r}
library(reshape2)
heat_df <- dcast(multi_results, phenotype ~ variable, value.var = "beta")
heat_matrix <- as.matrix(heat_df[,-1])
rownames(heat_matrix) <- heat_df$phenotype

pheatmap::pheatmap(heat_matrix, cluster_rows = TRUE, cluster_cols = TRUE,
                   main = "Beta Coefficients: Multivariable Model")

```


### Logistic Regression Model
```{r}
model_multi <- glm(GCA_present ~ age + sex + steroids_days + platelets + PMR, data = combined, family = binomial)
summary(model_multi)
```

```{r}
significant_results <- subset(multi_results, p_value < 0.05)
age_plot <- subset(multi_results, variable == "age")

ggplot(age_plot, aes(x = reorder(phenotype, beta), y = beta)) +
  geom_point(color = "steelblue") +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  coord_flip() +
  theme_minimal() +
  labs(title = "Adjusted Effect of Age on Histological Phenotypes",
       x = "Phenotype", y = "Beta Coefficient (95% CI)")
sex_plot <- subset(multi_results, variable == "sex")
steroids_plot <- subset(multi_results, variable == "steroids_days")

```

need to: 
- analyse results
- include RNA-seq data
