---
title: "Objective1"
output: html_document
---

```{r,  include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(ggplot2)
library(dplyr)
library(MASS)
library(scales)
library(broom)
```

```{r}
clin_var <- read.csv('/Users/kristinemedcalf/Desktop/Programming/Dissertation_GCA_Prediction/data/clinical_variables_cc_c1_c2.csv')
hist_var <- read.csv("/Users/kristinemedcalf/Desktop/Programming/Dissertation_GCA_Prediction/data/histological_variables_cc_c1_c2.csv")
combined <- merge(clin_var, hist_var, by = "ID")
```

Remove: 
• Any_granulomatous_infiltrate
• PALI
• Any_lymphocytic_infiltrate
• GCA_present
• Granulomatous_infiltrate_in_media
• Hyperplasia
• Fibrosis

```{r}
cols_to_remove <- c("Any_granulomatous_infiltrate", "PALI", "Any_lymphocytic_infiltrate", "GCA_present", "Granulomatous_infiltrate_in_media", "Hyperplasia", "Fibrosis")
hist_var <- hist_var[, setdiff(names(hist_var), cols_to_remove)]
trimmed_hist <- as.data.frame(hist_var[, -1])
```

MODEL TYPES:
Logistic Models: continuous or categorical features, binary outcomes
Linear Models: continuous (or encoded categorical) features, continuous outcomes
Classification Models: continuous or categorical features, categorical outcomes (2+ classes)

Variables:
Steroids_days → continuous
Sex → binary
Age (quartiles) → categorical (ordinal)

Modeling each outcome:
Binary features → Steroids_days: Linear model
Binary features → Sex: Logistic model
Binary features → Age (quartiles): Classification model
  Use multinomial logistic regression if unordered
  Use ordinal logistic regression if ordered


```{r}
binary_features <- c()
for (col in names(trimmed_hist)) {
  if (length(levels(factor(trimmed_hist[[col]]))) == 2) {
    binary_features <- c(binary_features, col)
  }
}

# Fit models
steroid_models <- lapply(binary_features, function(feature) {
  formula <- as.formula(paste("steroids_days ~", feature))
  lm(formula, data = combined)
})
names(steroid_models) <- binary_features

# Extract model results
linear_summary <- do.call(rbind, lapply(seq_along(steroid_models), function(i) {
  model <- steroid_models[[i]]
  feature <- binary_features[i]
  coef_summary <- summary(model)$coefficients
  
  if (feature %in% rownames(coef_summary)) {
    est <- coef_summary[feature, "Estimate"]
    se <- coef_summary[feature, "Std. Error"]
    p_val <- coef_summary[feature, "Pr(>|t|)"]
    ci_lower <- est - 1.96 * se
    ci_upper <- est + 1.96 * se
    data.frame(Feature = feature, Estimate = est, LowerCI = ci_lower, UpperCI = ci_upper, PValue = p_val)
  } else {
    data.frame(Feature = feature, Estimate = NA, LowerCI = NA, UpperCI = NA, PValue = NA)
  }
}))

# Filter valid rows and adjust p-values
linear_summary <- linear_summary[!is.na(linear_summary$Estimate), ]
linear_summary$AdjP <- p.adjust(linear_summary$PValue, method = "fdr")

# Forest plot
library(ggplot2)

ggplot(linear_summary, aes(x = reorder(Feature, Estimate), y = Estimate)) +
  geom_point() +
  geom_errorbar(aes(ymin = LowerCI, ymax = UpperCI), width = 0.2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  coord_flip() +
  labs(title = "Binary Features Predicting Steroid Days",
       x = "Histological Feature",
       y = "Estimated Effect") +
  theme_minimal()

```

Logistic Regression: Binary Features --> Sex (Binary)
```{r}
combined$sex <- factor(combined$sex, levels = c(1, 2), labels = c("Male", "Female"))

logistic_models <- vector("list", length(binary_features))
names(logistic_models) <- binary_features

for (i in seq_along(binary_features)) {
  feature <- binary_features[i]
  formula <- as.formula(paste("sex ~", feature))
  model_logistic <- glm(formula, data = combined, family = binomial)
  logistic_models[[i]] <- model_logistic
}

# Extract p-values
logistic_p_values <- sapply(seq_along(logistic_models), function(i) {
  model <- logistic_models[[i]]
  feature <- binary_features[i]
  coef_summary <- summary(model)$coefficients
  if (feature %in% rownames(coef_summary)) {
    return(coef_summary[feature, "Pr(>|z|)"])
  } else {
    return(NA)
  }
})

logistic_p_values <- p.adjust(logistic_p_values, method = "fdr")
names(logistic_p_values) <- binary_features


logistic_summary <- data.frame(
  Feature = character(),
  OR = numeric(),
  LowerCI = numeric(),
  UpperCI = numeric(),
  PValue = numeric(),
  stringsAsFactors = FALSE
)

for (i in seq_along(logistic_models)) {
  model <- logistic_models[[i]]
  feature <- binary_features[i]
  coef_summary <- summary(model)$coefficients
  
  if (feature %in% rownames(coef_summary)) {
    coef_val <- coef_summary[feature, "Estimate"]
    se <- coef_summary[feature, "Std. Error"]
    p_val <- coef_summary[feature, "Pr(>|z|)"]
    
    OR <- exp(coef_val)
    CI_lower <- exp(coef_val - 1.96 * se)
    CI_upper <- exp(coef_val + 1.96 * se)
    
    logistic_summary <- rbind(logistic_summary, data.frame(
      Feature = feature,
      OR = OR,
      LowerCI = CI_lower,
      UpperCI = CI_upper,
      PValue = p_val
    ))
  }
}

# Adjust p-values (FDR)
logistic_summary$AdjP <- p.adjust(logistic_summary$PValue, method = "fdr")

library(ggplot2)

ggplot(logistic_summary, aes(x = reorder(Feature, OR), y = OR)) +
  geom_point() +
  geom_errorbar(aes(ymin = LowerCI, ymax = UpperCI), width = 0.2) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
  coord_flip() +
  labs(title = "Odds Ratios for Binary Features Predicting Sex",
       x = "Feature",
       y = "Odds Ratio (Female vs Male)") +
  theme_minimal()

```
Ordinal Logistic Regression: 
```{r}
combined$age_quartile <- cut(
  combined$age,
  breaks = quantile(combined$age, probs = seq(0, 1, 0.25), na.rm = TRUE),
  include.lowest = TRUE,
  labels = c("Q1", "Q2", "Q3", "Q4"),
  ordered_result = TRUE
)

```

```{r}
# Make sure age_quartile is an ordered factor
combined$age_quartile <- factor(combined$age_quartile, ordered = TRUE)

# Load required package
library(MASS)

ordinal_models <- vector("list", length(binary_features))
names(ordinal_models) <- binary_features

for (i in seq_along(binary_features)) {
  feature <- binary_features[i]
  formula <- as.formula(paste("age_quartile ~", feature))
  model_ordinal <- polr(formula, data = combined, Hess = TRUE)
  ordinal_models[[i]] <- model_ordinal
}

# Extract p-values
ordinal_p_values <- sapply(seq_along(ordinal_models), function(i) {
  model <- ordinal_models[[i]]
  feature <- binary_features[i]
  coef_summary <- summary(model)$coefficients
  if (feature %in% rownames(coef_summary)) {
    zval <- coef_summary[feature, "Value"] / coef_summary[feature, "Std. Error"]
    return(2 * (1 - pnorm(abs(zval))))
  } else {
    return(NA)
  }
})

ordinal_p_values <- p.adjust(ordinal_p_values, method = "fdr")
names(ordinal_p_values) <- binary_features

library(MASS)

ordinal_summary <- data.frame(
  Feature = character(),
  OR = numeric(),
  LowerCI = numeric(),
  UpperCI = numeric(),
  PValue = numeric(),
  stringsAsFactors = FALSE
)

for (i in seq_along(ordinal_models)) {
  model <- ordinal_models[[i]]
  feature <- binary_features[i]
  coef_summary <- summary(model)$coefficients
  
  if (feature %in% rownames(coef_summary)) {
    estimate <- coef_summary[feature, "Value"]
    se <- coef_summary[feature, "Std. Error"]
    z <- estimate / se
    p_val <- 2 * (1 - pnorm(abs(z)))
    
    OR <- exp(estimate)
    CI_lower <- exp(estimate - 1.96 * se)
    CI_upper <- exp(estimate + 1.96 * se)
    
    ordinal_summary <- rbind(ordinal_summary, data.frame(
      Feature = feature,
      OR = OR,
      LowerCI = CI_lower,
      UpperCI = CI_upper,
      PValue = p_val
    ))
  }
}

# Adjust p-values for multiple testing
ordinal_summary$AdjP <- p.adjust(ordinal_summary$PValue, method = "fdr")

library(ggplot2)

ggplot(ordinal_summary, aes(x = reorder(Feature, OR), y = OR)) +
  geom_point() +
  geom_errorbar(aes(ymin = LowerCI, ymax = UpperCI), width = 0.2) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
  coord_flip() +
  labs(title = "Odds Ratios for Binary Features Predicting Age Quartile",
       x = "Histological Feature",
       y = "Odds Ratio (per increase in age quartile)") +
  theme_minimal()

```


```{r}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(ggplot2)
library(dplyr)
library(MASS)
library(scales)

### 1. Load and Clean Data
# Merge datasets
combined <- merge(clin_var, hist_var, by = "ID")

# Remove unwanted columns
cols_to_remove <- c("Any_granulomatous_infiltrate", "PALI", "Any_lymphocytic_infiltrate", 
                    "GCA_present", "Granulomatous_infiltrate_in_media", "Hyperplasia", "Fibrosis")
hist_var <- hist_var[, setdiff(names(hist_var), cols_to_remove)]

# Extract binary and categorical features
trimmed_hist <- hist_var[, -1]  # Drop ID column
binary_features <- c()
for (col in names(trimmed_hist)) {
  if (length(levels(factor(trimmed_hist[[col]]))) == 2) {
    binary_features <- c(binary_features, col)
  }
}
categorical_features <- setdiff(names(trimmed_hist), binary_features)

# Convert binary and categorical features to factors
for (col in c(binary_features, categorical_features)) {
  combined[[col]] <- factor(combined[[col]])
}

### 2. Linear Models: Predicting Steroid Days

all_features <- c(binary_features, categorical_features)
steroid_models <- lapply(all_features, function(feature) {
  lm(as.formula(paste("steroids_days ~", feature)), data = combined)
})
names(steroid_models) <- all_features

linear_summary <- do.call(rbind, lapply(seq_along(steroid_models), function(i) {
  model <- steroid_models[[i]]
  feature <- all_features[i]
  coef_summary <- summary(model)$coefficients
  rows <- rownames(coef_summary)
  rows <- rows[rows != "(Intercept)"]

  do.call(rbind, lapply(rows, function(rowname) {
    est <- coef_summary[rowname, "Estimate"]
    se <- coef_summary[rowname, "Std. Error"]
    p_val <- coef_summary[rowname, "Pr(>|t|)"]
    ci_lower <- est - 1.96 * se
    ci_upper <- est + 1.96 * se

    # Add level info only for categorical features
    is_categorical <- feature %in% categorical_features
    label <- if (is_categorical) paste0(rowname) else feature

    data.frame(
      Feature = label,
      Estimate = est,
      LowerCI = ci_lower,
      UpperCI = ci_upper,
      PValue = p_val
    )
  }))
}))

linear_summary$AdjP <- p.adjust(linear_summary$PValue, method = "fdr")

### 3. Logistic Models: Predicting Sex

combined$sex <- factor(combined$sex, levels = c(1, 2), labels = c("Male", "Female"))

logistic_models <- lapply(all_features, function(feature) {
  glm(as.formula(paste("sex ~", feature)), data = combined, family = binomial)
})
names(logistic_models) <- all_features

logistic_summary <- do.call(rbind, lapply(seq_along(logistic_models), function(i) {
  model <- logistic_models[[i]]
  feature <- all_features[i]
  coef_summary <- summary(model)$coefficients
  rows <- rownames(coef_summary)
  rows <- rows[rows != "(Intercept)"]

  do.call(rbind, lapply(rows, function(rowname) {
    est <- coef_summary[rowname, "Estimate"]
    se <- coef_summary[rowname, "Std. Error"]
    p_val <- coef_summary[rowname, "Pr(>|z|)"]
    or <- exp(est)
    ci_lower <- exp(est - 1.96 * se)
    ci_upper <- exp(est + 1.96 * se)

    # Conditionally label levels
    is_categorical <- feature %in% categorical_features
    label <- if (is_categorical) paste0(rowname) else feature

    data.frame(
      Feature = label,
      OR = or,
      LowerCI = ci_lower,
      UpperCI = ci_upper,
      PValue = p_val
    )
  }))
}))

logistic_summary$AdjP <- p.adjust(logistic_summary$PValue, method = "fdr")

### 4. Ordinal Logistic Models: Predicting Age Quartile

combined$age_quartile <- cut(
  combined$age,
  breaks = quantile(combined$age, probs = seq(0, 1, 0.25), na.rm = TRUE),
  include.lowest = TRUE,
  labels = c("Q1", "Q2", "Q3", "Q4"),
  ordered_result = TRUE
)

ordinal_models <- lapply(all_features, function(feature) {
  polr(as.formula(paste("age_quartile ~", feature)), data = combined, Hess = TRUE)
})
names(ordinal_models) <- all_features

ordinal_summary <- do.call(rbind, lapply(seq_along(ordinal_models), function(i) {
  model <- ordinal_models[[i]]
  feature <- all_features[i]
  coef_summary <- summary(model)$coefficients
  rows <- rownames(coef_summary)
  rows <- rows[rows != "(Intercept)"]
  do.call(rbind, lapply(rows, function(rowname) {
    est <- coef_summary[rowname, "Value"]
    se <- coef_summary[rowname, "Std. Error"]
    z_val <- est / se
    p_val <- 2 * (1 - pnorm(abs(z_val)))
    or <- exp(est)
    ci_lower <- exp(est - 1.96 * se)
    ci_upper <- exp(est + 1.96 * se)
    data.frame(Feature = paste0(feature, ":", rowname), OR = or, LowerCI = ci_lower, UpperCI = ci_upper, PValue = p_val)
  }))
}))
ordinal_summary$AdjP <- p.adjust(ordinal_summary$PValue, method = "fdr")

### 5. Forest Plots

# Increase margins for text
library(gridExtra)

# Linear model forest plot
ggplot(linear_summary, aes(x = reorder(Feature, Estimate), y = Estimate)) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = LowerCI, ymax = UpperCI), width = 0.2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  coord_flip() +
  labs(title = "Histological Predictors of Steroid Exposure Duration",
       x = "Feature",
       y = "Effect Estimate") +
  theme(plot.title = element_text(hjust = 0))  # Left-align title

# Logistic model forest plot
ggplot(logistic_summary, aes(x = reorder(Feature, OR), y = OR)) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = LowerCI, ymax = UpperCI), width = 0.2) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
  coord_flip() +
  labs(title = "Histological Predictors of Patient Sex",
       x = "Feature",
       y = "Odds Ratio") +
  theme(plot.title = element_text(hjust = 0))  # Left-align title

logistic_summary <- logistic_summary[logistic_summary$Feature != "Barcelona_score2", ]
ggplot(logistic_summary, aes(x = reorder(Feature, OR), y = OR)) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = LowerCI, ymax = UpperCI), width = 0.2) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
  coord_flip() +
  labs(title = "Histological Predictors of Patient Sex",
       x = "Feature",
       y = "Odds Ratio") +
  theme(plot.title = element_text(hjust = 0))  # Left-align title


# Ordinal logistic model forest plots

# Filter data for Q1|Q2
plot_q1_q2 <- ordinal_summary[grepl(":Q1\\|Q2$", ordinal_summary$Feature), ]

# Q1|Q2 plot
ggplot(plot_q1_q2, aes(x = reorder(Feature, OR), y = OR)) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = LowerCI, ymax = UpperCI), width = 0.2) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
  coord_flip() +
  labs(title = "Effect of Histological Features on Age Quartile Transition (Q1|Q2)",
       x = "Feature",
       y = "Odds Ratio") +
  theme(plot.title = element_text(hjust = 2))  # Left-align title

# Filter data for Q2|Q3
plot_q2_q3 <- ordinal_summary[grepl(":Q2\\|Q3$", ordinal_summary$Feature), ]

# Q2|Q3 plot
ggplot(plot_q2_q3, aes(x = reorder(Feature, OR), y = OR)) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = LowerCI, ymax = UpperCI), width = 0.2) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
  coord_flip() +
  labs(title = "Effect of Histological Features on Age Quartile Transition (Q2|Q3)",
       x = "Feature",
       y = "Odds Ratio") +
  theme(plot.title = element_text(hjust = 2))  # Left-align title

# Filter data for Q3|Q4
plot_q3_q4 <- ordinal_summary[grepl(":Q3\\|Q4$", ordinal_summary$Feature), ]

# Q3|Q4 plot
ggplot(plot_q3_q4, aes(x = reorder(Feature, OR), y = OR)) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = LowerCI, ymax = UpperCI), width = 0.2) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
  coord_flip() +
  labs(title = "Effect of Histological Features on Age Quartile Transition (Q3|Q4)",
       x = "Feature",
       y = "Odds Ratio") +
  theme(plot.title = element_text(hjust = 2))  # Left-align title

```

