---
title: "Confounding Analysis"
output: html_document
---
Loading clinical features and histological features.
```{r}
clin_var <- read.csv('/Users/kristinemedcalf/Desktop/Programming/Dissertation_GCA_Prediction/data/clinical_variables_cc_c1_c2.csv')
hist_var <- read.csv("/Users/kristinemedcalf/Desktop/Programming/Dissertation_GCA_Prediction/data/histological_variables_cc_c1_c2.csv")
combined <- merge(clin_var, hist_var, by = "ID")
```

Remove: 
• Any_granulomatous_infiltrate
• PALI
• Any_lymphocytic_infiltrate
• GCA_present

```{r}
cols_to_remove <- c("Any_granulomatous_infiltrate", "PALI", "Any_lymphocytic_infiltrate", "GCA_present")
hist_var <- hist_var[, setdiff(names(hist_var), cols_to_remove)]
```

Creating histograms for age and duration of steroids. Grouping them into 3-5 groups.

Age:
```{r}
par(mfrow = c(1, 2)) 

hist(combined$age, 
          col = "grey",
     main = "Default Histogram of Age",
     xlab = "Age",
     ylab = "Frequency")

quartiles <- round(quantile(combined$age, probs = seq(0, 1, 0.25), na.rm = TRUE))

hist(combined$age,
     breaks = quartiles,
     col = "grey",
     main = "Histogram of Age by Quartiles",
     xlab = "Age",
     ylab = "Frequency")

cat("Quartiles:\n")
quartiles

cat("\nGroupings:\n")
cat("Group 1:", quartiles[1], "-", quartiles[2]-1,"\n")
cat("Group 2:", quartiles[2], "-", quartiles[3]-1,"\n")
cat("Group 3:", quartiles[3], "-", quartiles[4]-1,"\n")
cat("Group 4:", quartiles[4], "-", quartiles[5],"\n")

```

Duration of Steroids:
```{r}
par(mfrow = c(1, 2)) 

hist(combined$steroids_days, 
     col = "grey",
     main = "Default Histogram ",
     xlab = "Steroid Duration (in days)",
     ylab = "Frequency")

summary(combined$steroids_days)
# outlier: 1833.00

steroid_trimmed <- combined$steroids_days[combined$steroids_days != 1833]

hist(steroid_trimmed, 
     col = "grey",
     main = "Trimmed Histogram",
     xlab = "Steroid Duration (in days)",
     ylab = "Frequency")

cat("\nGroupings:\n")
cat("Group 1: Less than 5 \n")
cat("Group 2: 5 - 10 \n")
cat("Group 3: More than 10 ")
```

Create a table of samples in each group for histological features. 
```{r}
trimmed_hist <- as.data.frame(hist_var[, -1])

for (col in names(trimmed_hist)) {
  if (is.numeric(trimmed_hist[[col]]) && length(unique(trimmed_hist[[col]])) < 10) {
    trimmed_hist[[col]] <- as.factor(trimmed_hist[[col]])
  }
}

grouped_cols <- sapply(trimmed_hist, function(x) is.factor(x) || is.character(x))
grouped_features <- names(trimmed_hist)[grouped_cols]

grouped_summary <- lapply(grouped_features, function(col) {
  tab <- table(trimmed_hist[[col]], useNA = "no") 
  data.frame(
    Feature = col,
    Level = names(tab),
    Count = as.integer(tab)
  )
})

hist_feature_summary <- do.call(rbind, grouped_summary)
```

Check the (sample) variance and select for low variance 
```{r}
variances <- sapply(trimmed_hist, function(x) {
  if (is.factor(x)) x <- as.numeric(x)
  var(x, na.rm = TRUE)
})

low_var_cols <- names(variances[variances < 0.05])
print(low_var_cols)

hist(variances, 
     main = "Histogram of Feature Variances",
     xlab = "Variance", 
     ylab = "Frequency",
     col = "lightblue", 
     breaks = 50, 
     xlim = c(0, 2))
```

Double-check using nearZeroVar()
```{r}
library(caret)
nzv <- nearZeroVar(trimmed_hist)
filtered_df <- df[, -nzv] 
```


Create a table of samples in each group for clinical features. 
```{r}
trimmed_clin <- as.data.frame(clin_var[, -1])

for (col in names(trimmed_clin)) {
  if (is.numeric(trimmed_clin[[col]]) && length(unique(trimmed_clin[[col]])) < 10) {
    trimmed_clin[[col]] <- as.factor(trimmed_clin[[col]])
  }
}

grouped_cols <- sapply(trimmed_clin, function(x) is.factor(x) || is.character(x))
grouped_features <- names(trimmed_clin)[grouped_cols]

grouped_summary <- lapply(grouped_features, function(col) {
  tab <- table(trimmed_clin[[col]], useNA = "no") 
  data.frame(
    Feature = col,
    Level = names(tab),
    Count = as.integer(tab)
  )
})

clin_feature_summary <- do.call(rbind, grouped_summary)
```


Correlation Plots

Phenotype:
```{r}
library(pheatmap)
phenotypes <- names(hist_var)[-1]
cor_matrix <- cor(combined[, phenotypes], use = "pairwise.complete.obs", method = "kendall") 
pheatmap(cor_matrix, cluster_rows = TRUE, cluster_cols = TRUE)
```

Clinical features:
```{r}
clinical_features <- names(clin_var)[-1]
numeric_clin <- combined[, clinical_features]
numeric_clin <- numeric_clin[, sapply(numeric_clin, is.numeric)]
numeric_clin <- numeric_clin[, apply(numeric_clin, 2, function(x) var(x, na.rm = TRUE) != 0)]
cor_matrix_clin <- cor(numeric_clin, use = "pairwise.complete.obs", method = "kendall") 
pheatmap(cor_matrix_clin, display_numbers = TRUE, cluster_rows = TRUE, cluster_cols = TRUE)
```

Use 2x2 frequency tables to analyze correlation between binary features? 
Use phi coeffecient as a measure of concordance 

```{r}
binary_features <- c()
for (col in names(trimmed_hist)) {
  if (length(levels(factor(trimmed_hist[[col]]))) == 2) {
    binary_features <- c(binary_features, col)
  }
}

phi_coefficient <- function(x, y) {
  tab <- table(x, y)
  if (!all(dim(tab) == c(2, 2))) {
    return(NA)
  }
  a <- tab[1,1]; b <- tab[1,2]; c <- tab[2,1]; d <- tab[2,2]
  num <- (a * d) - (b * c)
  denom <- sqrt((a + b) * (c + d) * (a + c) * (b + d))
  if (denom == 0) return(NA)
  return(num / denom)
}

n <- length(binary_features)
binaryfrequencytable <- matrix(NA, nrow = n, ncol = n)
colnames(binaryfrequencytable) <- binary_features
rownames(binaryfrequencytable) <- binary_features
diag(binaryfrequencytable) <- 1

df <- as.data.frame(lapply(trimmed_hist[binary_features], factor))
for (i in 1:(n - 1)) {
  for (j in (i + 1):n) {
    phi <- phi_coefficient(df[[i]], df[[j]])
    binaryfrequencytable[i, j] <- phi
    binaryfrequencytable[j, i] <- phi
  }
}

binaryfrequencytable <- as.data.frame(binaryfrequencytable)

```

phi coeffecient above 0.5 is a strong coeffecient
```{r}
col <- colnames(binaryfrequencytable)

for (i in 1:(length(col) - 1)) {
  for (j in (i + 1):length(col)) {
    val <- binaryfrequencytable[i, j]
    if (!is.na(val) && val > 0.8) {
      cat(paste("Feature 1:",col[i], "\nFeature 2:", col[j], "\nPhi = ", round(val, 3), "\n\n"))
    }
  }
}
```
Check missingness rate
```{r}
missing_rate <- colSums(is.na(trimmed_hist)) / nrow(trimmed_hist)
sort(missing_rate, decreasing = TRUE)

```

Remove: Granulomatous_infiltrate_in_media, Hyperplasia, Fibrosis
