---
title: "Confounding Analysis"
output: html_document
---
Loading clinical features and histological features.
```{r}
clin_var <- read.csv('/Users/kristinemedcalf/Desktop/Programming/Dissertation_GCA_Prediction/data/clinical_variables_cc_c1_c2.csv')
hist_var <- read.csv("/Users/kristinemedcalf/Desktop/Programming/Dissertation_GCA_Prediction/data/histological_variables_cc_c1_c2.csv")
names(hist_var)[names(hist_var) == "Lymphocytic_infiltrate_in._adventitia"] <- "Lymphocytic_infiltrate_in_adventitia"
combined <- merge(clin_var, hist_var, by = "ID")
```

Remove: 
• Any_granulomatous_infiltrate
• PALI
• Any_lymphocytic_infiltrate
• GCA_present

```{r}
cols_to_remove <- c("Any_granulomatous_infiltrate", "PALI", "Any_lymphocytic_infiltrate", "GCA_present")
hist_var <- hist_var[, setdiff(names(hist_var), cols_to_remove)]
```

Creating histograms for age and duration of steroids. Grouping them into 3-5 groups.

Age:
Based on quintiles
```{r}
quintiles <- round(quantile(combined$age, probs = seq(0, 1, 0.2), na.rm = TRUE))

combined$age_quintiles <- cut(combined$age,
                              breaks = quintiles,
                              include.lowest = TRUE,
                              right = TRUE)

table(combined$age_quintiles)

hist_result_quintiles <- hist(combined$age,
                              breaks = quintiles,
                              right = TRUE,
                              col = "steelblue",
                              main = "Histogram of Age by Quintiles",
                              xlab = "Age",
                              ylab = "Density",
                              freq = FALSE, 
                              ylim = c(0, 0.05))

text(x = hist_result_quintiles$mids,
     y = hist_result_quintiles$density,
     labels = hist_result_quintiles$counts,
     col = "black",
     pos = 3,
     cex = 0.9)
```

Based on quartiles.
```{r}
quartiles <- round(quantile(combined$age, probs = seq(0, 1, 0.25), na.rm = TRUE))

combined$age_quartiles <- cut(combined$age,
                          breaks = quartiles,
                          include.lowest = TRUE,
                          right = FALSE)

table(combined$age_quartiles)

hist_result_quartiles <- hist(combined$age,
     breaks = quartiles,
     right = FALSE,
     col = "steelblue",
     main = "Histogram of Age by Quartiles",
     xlab = "Age",
     ylab = "Density",
     freq = FALSE,
     ylim = c(0, 0.05))

text(x = hist_result_quartiles$mids,
     y = hist_result_quartiles$density,  
     labels = hist_result_quartiles$counts,
     col = "black",
     pos = 3,
     cex = 0.9)
```

Based on tertiles
```{r}
tertiles <- quantile(combined$age, probs = seq(0, 1, length.out = 4), na.rm = TRUE)

combined$age_tertile <- cut(combined$age,
                            breaks = tertiles,
                            include.lowest = TRUE,
                            right = FALSE)

table(combined$age_tertile)

hist_result_tertiles <- hist(combined$age,
     breaks = tertiles,
     right = FALSE,
     col = "steelblue",
     main = "Histogram of Age by Tertiles",
     xlab = "Age",
     ylab = "Density",
     freq = FALSE,
     ylim = c(0, 0.045))

text(x = hist_result_tertiles$mids,
     y = hist_result_tertiles$density,  
     labels = hist_result_tertiles$counts,
     col = "black",
     pos = 3,
     cex = 0.9)

```

Based on 11 year breaks: creates 3 groups with most even split
```{r}
hist(combined$age,main = "Histogram of Age", xlab = "Age", col = "steelblue")
table(combined$age)
min_age <- floor(min(combined$age, na.rm = TRUE))
max_age <- ceiling(max(combined$age, na.rm = TRUE))

eleven_year_breaks <- seq(min_age, max_age + 10, by = 10)

combined$age_with_eleven_year_breaks <- cut(combined$age,
                               breaks = eleven_year_breaks,
                               include.lowest = TRUE,
                               right = FALSE)

table(combined$age_with_eleven_year_breaks)

hist_result_eleven_years <- hist(combined$age,
     breaks = eleven_year_breaks,
     right = FALSE,
     col = "steelblue",
     main = "Histogram of Age Split by 11 Years",
     xlab = "Age",
     ylab = "Density",
     freq = FALSE, 
    ylim = c(0, 0.05))

text(x = hist_result_eleven_years$mids,
     y = hist_result_eleven_years$density,  
     labels = hist_result_eleven_years$counts,
     col = "black",
     pos = 3,
     cex = 0.9)
```

By eight year breaks: creates 4 groups with most even split
```{r}
eight_year_breaks <- seq(min_age, max_age, by = 8)

combined$age_with_eight_year_breaks <- cut(combined$age,
                               breaks = eight_year_breaks,
                               include.lowest = TRUE,
                               right = FALSE)

table(combined$age_with_eight_year_breaks)

hist_result_eight_years <- hist(combined$age,
     breaks = eight_year_breaks,
     right = FALSE,
     col = "steelblue",
     main = "Histogram of Age Split by 8 Years",
     xlab = "Age",
     ylab = "Density",
     freq = FALSE, 
      ylim = c(0, 0.06))

text(x = hist_result_eight_years$mids,
     y = hist_result_eight_years$density,  
     labels = hist_result_eight_years$counts,
     col = "black",
     pos = 3,
     cex = 0.9)
```

By 7 year breaks: creates 5 groups
```{r}
seven_year_breaks <- seq(min_age, max_age + 7, by = 7)

combined$age_with_seven_year_breaks <- cut(combined$age,
                               breaks = seven_year_breaks,
                               include.lowest = TRUE,
                               right = FALSE)

table(combined$age_with_seven_year_breaks)

hist_result_seven_years <- hist(combined$age,
     breaks = seven_year_breaks,
     right = FALSE,
     col = "steelblue",
     main = "Histogram of Age Split by 7 Years",
     xlab = "Age",
     ylab = "Density",
     freq = FALSE,
     ylim = c(0, 0.06))

text(x = hist_result_seven_years$mids,
     y = hist_result_seven_years$density,  
     labels = hist_result_seven_years$counts,
     col = "black",
     pos = 3,
     cex = 0.9)
```

Duration of Steroids:
```{r}
par(mfrow = c(1, 2)) 

hist(combined$steroids_days, 
     col = "grey",
     main = "Default Histogram ",
     xlab = "Steroid Duration (in days)",
     ylab = "Frequency")

summary(combined$steroids_days)
# outlier: 1833.00

steroid_trimmed <- combined$steroids_days[combined$steroids_days != 1833]

hist(steroid_trimmed, 
     col = "grey",
     main = "Trimmed Histogram",
     xlab = "Steroid Duration (in days)",
     ylab = "Frequency")

```

Clinical perspective
```{r}
breaks <- c(min(steroid_trimmed, na.rm = TRUE), 2, 10, max(steroid_trimmed, na.rm = TRUE) + 1)
breaks <- seq(min(steroid_trimmed, na.rm = TRUE), max(steroid_trimmed, na.rm = TRUE), by = 1)


hist_steroid_clinical <- hist(steroid_trimmed,
     breaks = breaks,
     col = "steelblue",
     main = "Histogram of Steroid Duration",
     xlab = "Steroid Duration (days)",
     ylab = "Density", 
     freq = FALSE, 
     ylim = c(0, 0.11))

text(x = hist_steroid_clinical$mids,
     y = hist_steroid_clinical$density,  
     labels = hist_steroid_clinical$counts,
     col = "black",
     pos = 3,
     cex = 0.9)
```

Non-clinical perspective
```{r}
tertiles <- quantile(steroid_trimmed, probs = seq(0, 1, length.out = 4), na.rm = TRUE)

hist_result_tertiles <- hist(steroid_trimmed,
     breaks = tertiles,
     right = FALSE,
     col = "steelblue",
     main = "Histogram of Steroid Exposure by Tertiles",
     xlab = "Steroid Exposure (in Days)",
     ylab = "Density",
     freq = FALSE,
     ylim = c(0, 0.12))

text(x = hist_result_tertiles$mids,
     y = hist_result_tertiles$density,  
     labels = hist_result_tertiles$counts,
     col = "black",
     pos = 3,
     cex = 0.9)
```

```{r}
quintiles <- round(quantile(steroid_trimmed, probs = seq(0, 1, 0.2), na.rm = TRUE))

hist_result_quintiles <- hist(steroid_trimmed,
                              breaks = quintiles,
                              right = TRUE,
                              col = "steelblue",
                              main = "Histogram of Steroid Exposure by Quintiles",
                              xlab = "Steroid Exposure (in Days)",
                              ylab = "Density",
                              freq = FALSE, 
                              ylim = c(0, 0.14))

text(x = hist_result_quintiles$mids,
     y = hist_result_quintiles$density,
     labels = hist_result_quintiles$counts,
     col = "black",
     pos = 3,
     cex = 0.9)

table(steroid_trimmed)
```

Based on quartiles.
```{r}
quartiles <- round(quantile(steroid_trimmed, probs = seq(0, 1, 0.25), na.rm = TRUE))

hist_result_quartiles <- hist(steroid_trimmed,
     breaks = quartiles,
     right = TRUE,
     col = "steelblue",
     main = "Histogram of Steroid Exposure by Quartiles",
     xlab = "Steroid Exposure (in Days)",
     ylab = "Density",
     freq = FALSE,
     ylim = c(0, 0.15))

text(x = hist_result_quartiles$mids,
     y = hist_result_quartiles$density,  
     labels = hist_result_quartiles$counts,
     col = "black",
     pos = 3,
     cex = 0.9)
```


Create a table of samples in each group for histological features. 
```{r}
library(tidyr)

trimmed_hist <- as.data.frame(hist_var[, -1])

for (col in names(trimmed_hist)) {
  if (is.numeric(trimmed_hist[[col]]) && length(unique(trimmed_hist[[col]])) < 10) {
    trimmed_hist[[col]] <- as.factor(trimmed_hist[[col]])
  }
}

grouped_cols <- sapply(trimmed_hist, function(x) is.factor(x) || is.character(x))
grouped_features <- names(trimmed_hist)[grouped_cols]

grouped_summary <- lapply(grouped_features, function(col) {
  tab <- table(trimmed_hist[[col]])
  df <- as.data.frame(tab)
  names(df) <- c("Level", "Count")
  df$Feature <- col
  return(df)
})

hist_feature_summary <- do.call(rbind, grouped_summary)
wide_hist_summary <- pivot_wider(hist_feature_summary,
                                  names_from = Level,
                                  values_from = Count,
                                  values_fill = NA)

wide_hist_summary <- wide_hist_summary[order(wide_hist_summary$Feature), ]

print(wide_hist_summary)

```

Check the (sample) variance and select for low variance 
```{r}
variances <- sapply(trimmed_hist, function(x) {
  if (is.factor(x)) x <- as.numeric(x)
  var(x, na.rm = TRUE)
})

low_var_cols <- names(variances[variances < 0.05])
print(low_var_cols)

hist(variances, 
     main = "Histogram of Feature Variances",
     xlab = "Variance", 
     ylab = "Frequency",
     col = "lightblue", 
     breaks = 50, 
     xlim = c(0, 2))

low_var_01 <- data.frame(sort(variances[variances < 0.1]))
names(low_var_01) <- c('Variance')
low_var_01
```

Double-check using nearZeroVar()
```{r}
library(caret)
nzv <- nearZeroVar(trimmed_hist)
filtered_df <- df[, -nzv] 
```

Create a table of samples in each group for clinical features. 
```{r}
trimmed_clin <- as.data.frame(clin_var[, -1])

for (col in names(trimmed_clin)) {
  if (is.numeric(trimmed_clin[[col]]) && length(unique(trimmed_clin[[col]])) < 10) {
    trimmed_clin[[col]] <- as.factor(trimmed_clin[[col]])
  }
}

grouped_cols <- sapply(trimmed_clin, function(x) is.factor(x) || is.character(x))
grouped_features <- names(trimmed_clin)[grouped_cols]

grouped_summary <- lapply(grouped_features, function(col) {
  tab <- table(trimmed_clin[[col]], useNA = "no") 
  data.frame(
    Feature = col,
    Level = names(tab),
    Count = as.integer(tab)
  )
})

clin_feature_summary <- do.call(rbind, grouped_summary)
```

Correlation Plots

Phenotype:
```{r}
library(pheatmap)
phenotypes <- names(hist_var)[-1]
cor_matrix <- cor(hist_var[, phenotypes], use = "pairwise.complete.obs")
rownames(cor_matrix) <- phenotypes
colnames(cor_matrix) <- phenotypes
pheatmap(cor_matrix,
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         cellheight = 8,
         fontsize_row = 7, 
         fontsize_col = 7)
```
*include all features in correlation plot

Clinical features:
```{r}
clinical_features <- names(clin_var)[-1]
numeric_clin <- combined[, clinical_features]
numeric_clin <- numeric_clin[, sapply(numeric_clin, is.numeric)]
numeric_clin <- numeric_clin[, apply(numeric_clin, 2, function(x) var(x, na.rm = TRUE) != 0)]
cor_matrix_clin <- cor(numeric_clin, use = "pairwise.complete.obs", method = "kendall") 
pheatmap(cor_matrix_clin, display_numbers = TRUE, cluster_rows = TRUE, cluster_cols = TRUE)
```

Use 2x2 frequency tables to analyze correlation between binary features? 
Use phi coeffecient as a measure of concordance 

```{r}
binary_features <- c()
for (col in names(trimmed_hist)) {
  if (length(levels(factor(trimmed_hist[[col]]))) == 2) {
    binary_features <- c(binary_features, col)
  }
}

phi_coefficient <- function(x, y) {
  tab <- table(x, y)
  if (!all(dim(tab) == c(2, 2))) {
    return(NA)
  }
  a <- tab[1,1]; b <- tab[1,2]; c <- tab[2,1]; d <- tab[2,2]
  num <- (a * d) - (b * c)
  denom <- sqrt((a + b) * (c + d) * (a + c) * (b + d))
  if (denom == 0) return(NA)
  return(num / denom)
}

n <- length(binary_features)
binaryfrequencytable <- matrix(NA, nrow = n, ncol = n)
colnames(binaryfrequencytable) <- binary_features
rownames(binaryfrequencytable) <- binary_features
diag(binaryfrequencytable) <- 1

df <- as.data.frame(lapply(trimmed_hist[binary_features], factor))
for (i in 1:(n - 1)) {
  for (j in (i + 1):n) {
    phi <- phi_coefficient(df[[i]], df[[j]])
    binaryfrequencytable[i, j] <- phi
    binaryfrequencytable[j, i] <- phi
  }
}

binaryfrequencytable <- as.data.frame(binaryfrequencytable)

```

phi coeffecient above 0.5 is a strong coefficient
```{r}
col <- colnames(binaryfrequencytable)
remove_corr <- data.frame(Feature1 = character(), Feature2 = character(), Correlation = numeric(), stringsAsFactors = FALSE)

for (i in 1:(length(col) - 1)) {
for (j in (i + 1):length(col)) {
 val <- binaryfrequencytable[i, j]
 if (!is.na(val) && val > 0.8) {
   remove_corr <- rbind(remove_corr, data.frame(
     Feature1 = col[i],
     Feature2 = col[j],
     Correlation = round(val, 3)
   ))
   cat(paste("Feature 1:", col[i], "\nFeature 2:", col[j], "\nPhi =", round(val, 3), "\n\n"))
 }
}
}

remove_corr
```

Check missingness rate
```{r}
missing_rate <- colSums(is.na(trimmed_hist)) / nrow(trimmed_hist)
sort(missing_rate, decreasing = TRUE)

```

Remove: Granulomatous_infiltrate_in_media, Hyperplasia, Fibrosis

Cross-Tabulation 
```{r}
library(writexl)
infiltrates <- data.frame(
  Adventitia = combined$Granulomatous_infiltrate_in_adventitia,
  Intima     = combined$Granulomatous_infiltrate_in_intima,
  Media      = combined$Granulomatous_infiltrate_in_media,
  Aggregates = combined$Aggregates
)
#write_xlsx(infiltrates, "infiltrates.xlsx")

patterns <- data.frame(
  Adventitia = combined$Media_pattern,
  Intima     = combined$Intima_pattern,
  Media      = combined$Adventitia_pattern)

# write_xlsx(patterns, "patterns.xlsx")

lymph <- data.frame(
  Adventitia = combined$Lymphocytic_infiltrate_in_adventitia,
  Intima     = combined$Lymphocytic_infiltrate_in_intima,
  Media      = combined$Lymphocytic_infiltrate_in_media, 
  Hyperplasia = combined$Hyperplasia,
  Fibrosis = combined$Fibrosis)

# write_xlsx(lymph, "lymph.xlsx")


others <- data.frame(
  Adventitia = combined$Lymphocytic_infiltrate_in_adventitia,
  Intima     = combined$Lymphocytic_infiltrate_in_intima,
  AdventitiaPattern      = combined$Adventitia_pattern, 
  IntimaPattern      = combined$Intima_pattern, 
  BarcelonaScore      = combined$Barcelona_score, 
  OcclusionGrade      = combined$Occlusion_grade, 
  MediaDestruction      = combined$Media_destruction)

# write_xlsx(others, "others.xlsx")

```
